---
title: "Add HRL Habitat Data"
author: "Liz Stebbins (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.dim = c(8, 6)
)
```

```{r setup, include = FALSE}
library(DSMhabitat)
library(tidyverse)
library(lubridate)
library(DSMflow)
library(readxl)
```

# Overview

As part of a R2R scenario, habitat commitments made under the Healthy Rivers and Landscapes (HRL) project were included for the following watersheds: 

* American River
* Feather River
* Mokelumne River
* Yuba River
* Tuolumne River

Staff from the Tuolumne River and Yuba River provided data directly for modeling anticipated flow-to-suitable area relationships under the HRL commitments. These streams developed methods to scale an existing habitat flow-to-suitable area relationship based on existing and proposed habitat projects and can be read about in more detail *here* (TODO).

The Mokelumne, American, and Feather Rivers did not provided updated relationships, and for consistency across the basin R2R used the flow-to-suitable area relationships provided for the [Scientific Basis Report (SBR), Section 6](https://www.waterboards.ca.gov/waterrights/water_issues/programs/bay_delta/docs/2023/staff-report/app-g2.pdf). 

# Methods

For use in the R2R scenario, we implemented the following methods:

1) Read in the flow-to-suitable area relationship used in the SBR
2) Functionalize the relationship 
3) Generate estimated habitat using flows in the HRL CalSim3 run for the model years (`1980-2000`)

# American River

## HRL flow-to-habitat relationship

Instream and floodplain rearing habitat were provided. They are the same as those used in the SBR.

**Data Source:** FlowWest coordinated with representatives for the American River to determine the best scientific information available to quantify existing suitable habitat, and to describe the VA non-flow habitat as *suitable area that varies with flow*. Habitat data were provided to R2R for use in accounting for planned habitat projects by [Erica Bishop](mailto:ebishop@waterforum.org) and [Chris Hammersmark](mailto:c.hammersmark@cbecoeng.com). Data provided represented the HRL commitments based on flow-to-habitat relationships borne out by recent EIP designs and constructed projects for the subject habitat type. 

```{r, echo=FALSE}
american_hrl <- read_excel("data-raw/R2R_HRL_habitat_inputs/data/HRL_habitat_commitments_for_R2R_American.xlsx",
                               range = "A1:G10") |> 
  janitor::clean_names() |>
  mutate(duration_inundated_days = ifelse(duration_inundated_days == "N/A", NA, duration_inundated_days),
         water_year_type = ifelse(water_year_type == "N/A", NA, water_year_type),
         run = ifelse(run == "fall run", "fall", run)) |> 
  mutate(flow_cfs = as.numeric(flow_to_inundate_suitable_acres),
         flow_cfs = ifelse(is.na(flow_cfs), 20000, flow_cfs), 
         scenario = "HRL", 
         watershed = "American River",
         FR_juv_sqm_hrl = DSMhabitat::acres_to_square_meters(suitable_acres)) |> 
  select(watershed, flow_cfs, FR_juv_sqm_hrl, scenario)

knitr::kable(head(american_hrl))
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
existing_rearing <- DSMhabitat::american_river_instream |>
  filter(!is.na(FR_juv_sqm)) |> 
  select(flow_cfs, FR_juv_sqm)

existing_floodplain <- DSMhabitat::american_river_floodplain |> 
  filter(!is.na(FR_floodplain_acres)) |> 
  select(flow_cfs, FR_floodplain_acres)

existing_rearing_fn <- approxfun(existing_rearing$flow_cfs, existing_rearing$FR_juv_sqm, rule = 2)

existing_fp_fn <- approxfun(existing_floodplain$flow_cfs, existing_floodplain$FR_floodplain_acres, rule = 2)

existing_and_HRL <- american_hrl |> 
  mutate(existing_rear = existing_rearing_fn(flow_cfs),
         existing_fp = existing_fp_fn(flow_cfs),
         fp_HRL = ifelse(flow_cfs <= 5200, NA, DSMhabitat::square_meters_to_acres(FR_juv_sqm_hrl)),
         existing_and_HRL = ifelse(flow_cfs <= 5200, existing_rear + FR_juv_sqm_hrl, existing_fp + fp_HRL)) |> 
  select(-scenario)

# juvenile
existing_and_HRL |> 
  mutate(type = ifelse(flow_cfs <= 5200, "juv", "fp"),
         existing = ifelse(flow_cfs <= 5200, existing_rear, existing_fp)) |> 
  pivot_longer(c(existing, existing_and_HRL),
               names_to = "scenario",
               values_to = "value") |> 
  ggplot(aes(x = flow_cfs, y = value, color = scenario)) + 
  geom_line() +
  facet_wrap(~type, scales = "free") +
  geom_vline(aes(xintercept = 5200), linetype = "dashed") +
  theme_minimal()
```

# Feather River

## HRL flow-to-habitat relationship

Spawning and instream rearing actions are only for the LFC.

**Data Source:** Data came from the FERC relicensing spawning WUA/RSI and were provided for the SBR by Jason Kindopp. 

```{r, echo=FALSE}
feather_hrl <- read_csv(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", "feather_flow_habitat.csv")) |> 
  filter(scenario == "VA") |> 
  mutate(habitat_area_sqm = DSMhabitat::acres_to_square_meters(habitat_area_acres),
         scenario = "HRL") |> 
  select(watershed, location, flow_cfs, habitat_type, habitat_area_sqm) |> 
  pivot_wider(id_cols = c(watershed, location, flow_cfs), 
              names_from = habitat_type,
              values_from = habitat_area_sqm) |> 
  rename(FR_spw_sqm_hrl = Spawning,
         FR_juv_ic_sqm_hrl = `Instream rearing`,
         FR_juv_fp_sqm_hrl = `Floodplain rearing`) |> 
  glimpse()

knitr::kable(head(feather_hrl))
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
feather_hrl |> 
  pivot_longer(FR_spw_sqm_hrl:FR_juv_fp_sqm_hrl,
               names_to = "lifestage",
               values_to = "value") |> 
  ggplot(aes(x = flow_cfs, y = value, color = lifestage)) + 
  geom_line() +
  facet_wrap(~location, scales = "free") +
  theme_minimal()
```

# Mokelumne River

## HRL flow-to-habitat relationship

Instream and floodplain rearing habitat were provided. They are the same as those used in the SBR.

**Data Source:** Data were provided by Robyn Bilski for the SBR.

```{r, echo=FALSE}
spawning <- read_excel('data-raw/R2R_HRL_habitat_inputs/data/moke_flow_to_habitat_RB_edits.xlsx', 
                           sheet = 'spawning', 
                           range = 'A1:D5') |> 
  mutate(FR_sqm_hrl = DSMhabitat::acres_to_square_meters(FR_spawn_acres_VA_enhanced),
         lifestage = "Spawning") |> 
  select(flow_cfs, FR_sqm_hrl, lifestage)

rearing <- read_excel('data-raw/R2R_HRL_habitat_inputs/data/moke_flow_to_habitat_RB_edits.xlsx', 
                      sheet = 'instream_rearing', 
                      range = 'A1:G17') |> 
  mutate(FR_juv_acres_VAadditional = 0.5*(FR_fry_acres_VA_enhanced),
         FR_fry_acres_VA_enhanced = 0.5*(FR_fry_acres_VA_enhanced) + FR_fry_acres_existing,
         FR_juv_acres_VA_enhanced = 0.5*(FR_juv_acres_VA_enhanced) + FR_juv_acres_existing) |> 
  mutate(FR_juv_sqm_hrl = acres_to_square_meters(FR_juv_acres_VA_enhanced),
         FR_fry_sqm_hrl = acres_to_square_meters(FR_fry_acres_VA_enhanced)) |> 
  pivot_longer(FR_juv_sqm_hrl:FR_fry_sqm_hrl,
               names_to = "lifestage",
               values_to = "FR_sqm_hrl") |> 
  select(flow_cfs, FR_sqm_hrl, lifestage) 

mokelumne_hrl <- bind_rows(spawning, rearing) |> 
  mutate(scenario = "HRL",
         watershed = "Mokelumne") |> 
  pivot_wider(id_cols = flow_cfs,
              names_from = lifestage,
              values_from = FR_sqm_hrl) |> 
  rename(FR_spw_sqm_hrl = Spawning)

knitr::kable(head(mokelumne_hrl))
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
mokelumne_hrl |> 
  pivot_longer(FR_spw_sqm_hrl:FR_fry_sqm_hrl,
               names_to = "lifestage",
               values_to = "value") |> 
  ggplot(aes(x = flow_cfs, y = value)) + 
  geom_line() +
  facet_wrap(~lifestage, scales = "free") +
  theme_minimal()
```


# Add to data objects

* functionalize curves
* use LTO_12a flows

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# TODO check these methods. do we need to do something more nuanced? probably
# american
american_juv_fn <- approxfun(american_hrl$flow_cfs, spawning$FR_juv_sqm_hrl,
                            rule = 2)

# feather
feather_spw_fn <- approxfun(feather_hrl$flow_cfs, feather_hrl$FR_spw_sqm_hrl,
                            rule = 2)
feather_juv_fn <- approxfun(feather_hrl$flow_cfs, feather_hrl$FR_juv_ic_sqm_hrl,
                            rule = 2)
feather_fp_fn <- approxfun(feather_hrl$flow_cfs, feather_hrl$FR_juv_fp_sqm_hrl,
                            rule = 2)

# mokelumne
mokelumne_spw_fn <- approxfun(mokelumne_hrl$flow_cfs,
                              mokelumne_hrl$FR_spw_sqm_hrl,
                              rule = 2)
mokelumne_juv_fn <- approxfun(mokelumne_hrl$flow_cfs,
                              mokelumne_hrl$FR_juv_sqm_hrl,
                              rule = 2)
# TODO not doing fry for mokelumne

# tuolumne
tuolumne_hrl <- read_csv(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", "tuolumne_hrl.csv")) |> 
  mutate(fp_sqm_va = DSMhabitat::acres_to_square_meters(fp_acres_va))

tuolumne_spw_fn <- approxfun(tuolumne_hrl$flow_cfs,
                             tuolumne_hrl$fr_spw_sqm_va,
                             rule = 2)
tuolumne_juv_fn <- approxfun(tuolumne_hrl$flow_cfs,
                             tuolumne_hrl$fr_juv_sqm_va,
                             rule = 2)
tuolumne_fp_fn <- approxfun(tuolumne_hrl$flow_cfs,
                             tuolumne_hrl$fp_sqm_va,
                             rule = 2)

# yuba
yuba_hrl <- read_csv(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", "yuba_hrl.csv"))

yuba_juv_fn <- approxfun(tuolumne_hrl$flow_cfs,
                         tuolumne_hrl$fr_juv_ic_sqm,
                         rule = 2)
yuba_fp_fn <- approxfun(tuolumne_hrl$flow_cfs,
                        tuolumne_hrl$fr_juv_fp_sqm,
                        rule = 2)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
flows <- DSMflow::flows_cfs$LTO_12a |>
  filter(year(date) >= 1980 & year(date) <= 2000) |> 
  pivot_longer(`Antelope Creek`:`San Joaquin River`,
               names_to = "watershed",
               values_to = "flow_cfs") |> 
  mutate(month = month(date),
         year = year(date)) |> 
  group_by(year, month, watershed) |> 
  summarise(median_flow_cfs = median(flow_cfs)) |> 
  ungroup()

update_hrl_habitat_from_SBR <- function(flows, watershed_name, hab_fn) {
  new_hab <- flows |> 
    filter(watershed == watershed_name) |> 
    mutate(new_hab = hab_fn(median_flow_cfs)) |> 
    pivot_wider(id_cols = month, names_from = year, values_from = median_flow_cfs) |>
    select(-month) |> 
    as.matrix()
  
  rownames(new_hab) <- month.abb
  
  return(new_hab)
  
}

# update spawning
# TODO fix this, producing very low habitat
DSMhabitat::fr_spawn$r_to_r_hrl["Feather River", , ] <- update_hrl_habitat_from_SBR(flows, "Feather River",
                                                                                    feather_spw_fn)
DSMhabitat::fr_spawn$r_to_r_hrl["Mokelumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Mokelumne River",
                                                                                    mokelumne_spw_fn)
DSMhabitat::fr_spawn$r_to_r_hrl["Tuolumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Tuolumne River",
                                                                                    tuolumne_spw_fn)
# update ic rearing
DSMhabitat::fr_juv$r_to_r_hrl["American River", , ] <- update_hrl_habitat_from_SBR(flows, "American River",
                                                                                    american_juv_fn)
DSMhabitat::fr_juv$r_to_r_hrl["Feather River", , ] <- update_hrl_habitat_from_SBR(flows, "Feather River",
                                                                                    feather_juv_fn)
DSMhabitat::fr_juv$r_to_r_hrl["Mokelumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Mokelumne River",
                                                                                    mokelumne_juv_fn)
DSMhabitat::fr_juv$r_to_r_hrl["Tuolumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Tuolumne River",
                                                                                    tuolumne_juv_fn)
DSMhabitat::fr_juv$r_to_r_hrl["Yuba River", , ] <- update_hrl_habitat_from_SBR(flows, "Yuba River",
                                                                                    feather_juv_fn)

# update fp rearing
DSMhabitat::fr_fp$r_to_r_hrl["Feather River", , ] <- update_hrl_habitat_from_SBR(flows, "Feather River",
                                                                                    feather_fp_fn)
DSMhabitat::fr_fp$r_to_r_hrl["Tuolumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Tuolumne River",
                                                                                    tuolumne_fp_fn)
DSMhabitat::fr_fp$r_to_r_hrl["Yuba River", , ] <- update_hrl_habitat_from_SBR(flows, "Yuba River",
                                                                                    feather_fp_fn)

# TODO add this script to update_data.R so it's run 
```
