---
title: "Add HRL Habitat Data"
author: "Liz Stebbins (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.dim = c(8, 6)
)
```

```{r setup, include = FALSE}
library(DSMhabitat)
library(tidyverse)
library(lubridate)
library(DSMflow)
library(readxl)
library(sf)
#remotes::install_github("flowwest/riceflows4ff")
library(riceflows4ff)
library(scales)

plot_palette <-c("Existing" = "#1B9E77", 
                 "HRL" = "#D95F02")

```

# Overview

Habitat and flow commitments made under the Healthy Rivers and Landscapes (HRL) program were included as inputs to the R2R modeling framework for the following watersheds:

-   American River
-   Feather River
-   Mokelumne River
-   Yuba River
-   Tuolumne River

Staff from the Tuolumne River and Yuba River provided flow-to-suitable area relationships representing anticipated effects of the HRL commitments. These watersheds developed methodology to scale existing flow-to-suitable habitat relationships based on existing and proposed habitat projects. Documentation for the Tuolumne River methodology is available [here](https://va-hrl-docs.s3.us-east-1.amazonaws.com/tuolumne/Tuolumne+River+VA+Habitat+Modeling+for+SBR+final+revision+20240802.pdf) and for the Yuba River methodology [here](https://va-hrl-docs.s3.us-east-1.amazonaws.com/yuba/Summary+of+LYR+Hydraulic+Modeling+Methodology_7-1-2024.docx). 

For the Mokelumne, American, and Feather Rivers, R2R used the flow-to-suitable area relationships provided for the [Scientific Basis Report (SBR), Section 6](https://www.waterboards.ca.gov/waterrights/water_issues/programs/bay_delta/docs/2023/staff-report/app-g2.pdf) to ensure consistency across the basin.

# Methods

The following workflow was used to incorporate these relationships into the R2R habitat model framework: 

1)  Read in the flow-to-suitable area relationship used in the SBR.
2)  Functionalize the relationship using interpolation with flow as the predictive variable and suitable habitat as the response variable.
3)  Generate estimated habitat using flows from the HRL CalSim3 run for the model years (`1980-2000`).

# Read and functionalize flow-to-area curves

## American River

### HRL flow-to-habitat relationship

Instream and floodplain rearing habitat were provided. They are the same as those used in the SBR.

**Data Source:** FlowWest coordinated with representatives for the American River to determine the best scientific information available to quantify existing suitable habitat, and to describe the VA non-flow habitat as suitable area that varies with flow. Habitat data were provided to R2R for use in accounting for planned habitat projects by [Erica Bishop](mailto:ebishop@waterforum.org) and [Chris Hammersmark](mailto:c.hammersmark@cbecoeng.com). Data provided represented the HRL commitments based on flow-to-habitat relationships borne out by recent EIP designs and constructed projects for the subject habitat type.

```{r, echo=FALSE, message = FALSE, warning = FALSE}
american_hrl <- read_excel("data-raw/R2R_HRL_habitat_inputs/data/HRL_habitat_commitments_for_R2R_American.xlsx",
                               range = "A1:G10") |> 
  janitor::clean_names() |>
  mutate(duration_inundated_days = ifelse(duration_inundated_days == "N/A", NA, duration_inundated_days),
         water_year_type = ifelse(water_year_type == "N/A", NA, water_year_type),
         run = ifelse(run == "fall run", "fall", run)) |> 
  mutate(flow_cfs = as.numeric(flow_to_inundate_suitable_acres),
         flow_cfs = ifelse(is.na(flow_cfs), 20000, flow_cfs), 
         scenario = "HRL", 
         watershed = "American River",
         FR_juv_sqm_hrl = DSMhabitat::acres_to_square_meters(suitable_acres))

knitr::kable(head(american_hrl |> 
                    mutate(FR_juv_sqm_hrl = prettyNum(square_meters_to_acres(FR_juv_sqm_hrl), big.mark = ",")) |> 
                    select(Watershed = watershed, 
                           `Flow (cfs)` = flow_cfs, 
                           `VA juvenile habitat (acres)` = FR_juv_sqm_hrl)))
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
existing_rearing <- DSMhabitat::american_river_instream |>
  filter(!is.na(FR_juv_sqm)) |> 
  select(flow_cfs, FR_juv_sqm)

existing_floodplain <- DSMhabitat::american_river_floodplain |> 
  filter(!is.na(FR_floodplain_acres)) |> 
  select(flow_cfs, FR_floodplain_acres)

existing_rearing_fn <- approxfun(existing_rearing$flow_cfs, existing_rearing$FR_juv_sqm, rule = 2)

existing_fp_fn <- approxfun(existing_floodplain$flow_cfs, existing_floodplain$FR_floodplain_acres, rule = 2)

existing_and_HRL <- american_hrl |> 
  mutate(existing_rear = existing_rearing_fn(flow_cfs),
         existing_fp = existing_fp_fn(flow_cfs),
         fp_HRL = ifelse(flow_cfs <= 5200, NA, DSMhabitat::square_meters_to_acres(FR_juv_sqm_hrl)),
         existing_and_HRL = ifelse(flow_cfs <= 5200, existing_rear + FR_juv_sqm_hrl, existing_fp + fp_HRL)) |> 
  select(-scenario)

# juvenile
existing_and_HRL |> 
  mutate(type = ifelse(flow_cfs <= 5200, "juv", "fp"),
         existing = ifelse(flow_cfs <= 5200, existing_rear, existing_fp)) |> 
  pivot_longer(c(existing, existing_and_HRL),
               names_to = "scenario",
               values_to = "value") |> 
  mutate(scenario = ifelse(scenario == "existing", "Existing", "HRL"),
         type = ifelse(type == "fp", "Floodplain", 
                       "Juvenile")) |> 
  ggplot(aes(x = flow_cfs, y = value, color = scenario)) + 
  geom_line() +
  facet_wrap(~type, scales = "free") +
  geom_vline(aes(xintercept = 5200), linetype = "dashed") +
  theme_minimal() +
  scale_color_manual(values = plot_palette) +
  labs(x = "Flow (cfs)",
       y = "Suitable acres")  +
  theme(legend.justification = c(1,0), legend.position = c(.9,.3))
```

## Feather River

### HRL flow-to-habitat relationship

Spawning and instream rearing actions are only for the LFC.

**Data Source:** Data came from the FERC relicensing spawning WUA/RSI and were provided for the SBR by Jason Kindopp.

```{r, echo=FALSE, message = FALSE, warning = FALSE}
# code ported directly from va-habitat/habitat-functions/Feather/feather.Rmd

calsim <- read_excel(here::here("data-raw", "R2R_HRL_habitat_inputs", "data",
                                'Feather_LFC_pulse_flow_DRAFT_051619.xlsm'), 
                     sheet = 'Output')

# existing spawning
upper_spawning <- read_excel(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", 'Feathe_River_Upper_and_Lower_Salmon_and_Steelhead_Spawning_RSI_for_FERC_relicensing.xlsx'),
                             sheet = 'Upper', range = 'A2:B32') |> 
  mutate(section = 'LFC')

lower_spawning <- read_excel(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", 'Feathe_River_Upper_and_Lower_Salmon_and_Steelhead_Spawning_RSI_for_FERC_relicensing.xlsx'),
                             sheet = 'Lower', range = 'A2:B32') |> 
  mutate(section = 'HFC')

feather_spawn_length <- DSMhabitat::watershed_lengths |> 
  filter(watershed == 'Feather River', species == 'fr', lifestage == 'spawning') |> 
  pull(feet)

lfc_spawn_length <- 42240
hfc_spawn_length <- feather_spawn_length - lfc_spawn_length

lfc_spawning <- upper_spawning |> 
  mutate(suitable_acres = round(`Chinook RSI`/1000*lfc_spawn_length/43560, 2)) |> 
  select(flow_cfs = `Flow (cfs)`, suitable_acres, section)

hfc_spawning <- lower_spawning |> 
  mutate(suitable_acres = round(`Chinook RSI`/1000*hfc_spawn_length/43560, 2)) |> 
  select(flow_cfs = `Flow (cfs)`, suitable_acres, section)

existing_spawn_lfc <- approxfun(lfc_spawning$flow_cfs, lfc_spawning$suitable_acres,
                                rule = 2)

existing_spawn_hfc <- approxfun(hfc_spawning$flow_cfs, hfc_spawning$suitable_acres,
                                rule = 2)

# existing ic rearing
upper_fry <- read_excel(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", 'fry_juv_rsi_addendum_2.xlsx'), sheet = 'upper_fry')
lower_fry <- read_excel(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", 'fry_juv_rsi_addendum_2.xlsx'), sheet = 'lower_fry')
upper_juv <- read_excel(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", 'fry_juv_rsi_addendum_2.xlsx'), sheet = 'upper_juv')
lower_juv <- read_excel(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", 'fry_juv_rsi_addendum_2.xlsx'), sheet = 'lower_juv')


feather_length <- DSMhabitat::watershed_lengths |> 
  filter(watershed == 'Feather River', species == 'fr', lifestage == 'rearing') |> 
  pull(feet)

# refactor to use just one variable 
lfc_length <- 42240
hfc_length <- feather_length - lfc_length

# use low no cover instead of 0.0 no cover
rearing_lfc <- bind_rows(upper_fry, upper_juv) |> 
  filter(cover != 'No Cover 0.0') |> 
  select(-cover) |> 
  mutate(suitable_acres = round(rsi/1000*lfc_length/43560, 2), flow_cfs = round(flow_cfs)) |>
  select(flow_cfs, suitable_acres, lifestage, section)

rearing_hfc <- bind_rows(lower_fry, lower_juv) |> 
  filter(cover != 'No Cover 0.0') |> 
  select(-cover) |> 
  mutate(suitable_acres = round(rsi/1000*hfc_length/43560, 2), flow_cfs = round(flow_cfs)) |>
  select(flow_cfs, suitable_acres, lifestage, section)

rearing_hfc_fry <- rearing_hfc |> 
  filter(lifestage == 'fry')

rearing_hfc_juv <- rearing_hfc |> 
  filter(lifestage == 'juvenile')

rearing_lfc_fry <- rearing_lfc |> 
  filter(lifestage == 'fry')

rearing_lfc_juv <- rearing_lfc |> 
  filter(lifestage == 'juvenile')

existing_rearing_hfc_fry <- approxfun(rearing_hfc_fry$flow_cfs, rearing_hfc_fry$suitable_acres,
                                      rule = 2)
existing_rearing_hfc_juv <- approxfun(rearing_hfc_juv$flow_cfs, rearing_hfc_juv$suitable_acres,
                                      rule = 2)

existing_rearing_lfc_fry <- approxfun(rearing_lfc_fry$flow_cfs, rearing_lfc_fry$suitable_acres,
                                      rule = 2)
existing_rearing_lfc_juv <- approxfun(rearing_lfc_juv$flow_cfs, rearing_lfc_juv$suitable_acres,
                                      rule = 2)

# existing fp
existing_fp <- DSMhabitat::feather_river_floodplain |> 
  mutate(suitable_acres = FR_floodplain_acres * .27) |> 
  select(flow_cfs, suitable_acres)

existing_fp_hab <- approxfun(existing_fp$flow_cfs, existing_fp$suitable_acres, rule = 2)

# VA 
va_proj <- read_excel(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", 'VA_Habitat_value_of_FR_projects_updated_5-9-19.xlsx'), 
                      sheet = 'sadie', range = 'A1:L12')

# VA spawn
new_spawning <- va_proj |>
  filter(habitat_type == 'spawning') |>
  select(reach:max_flow_suitability) |>
  mutate(acres = 3.75 + 11.6) |>
  unique()

new_spawning_acres <- approxfun(x = c(new_spawning$min_flow, new_spawning$target_flow, new_spawning$max_flow),
                                y = c(new_spawning$acres * new_spawning$min_flow_suitability, 
                                      new_spawning$acres * new_spawning$target_flow_suitability,
                                      new_spawning$acres * new_spawning$max_flow_suitability),
                                yleft = 0, rule = 2)

va_enhanced_spawn_lfc <- tibble(
  flow_cfs = lfc_spawning$flow_cfs,
  suitable_acres = existing_spawn_lfc(flow_cfs) + new_spawning_acres(flow_cfs)
)

bind_rows(
  mutate(lfc_spawning, scenario = 'Existing'),
  mutate(va_enhanced_spawn_lfc, scenario = 'HRL')) |> 
  ggplot(aes(flow_cfs, suitable_acres, color = scenario)) +
  geom_line() +
  theme_minimal() +
  labs(y = 'Suitable acres', x = 'Flow (cfs)', title = "Spawning (LFC)") +
  scale_color_manual(values = plot_palette) +
  theme(legend.justification = c(1,0), legend.position = c(.9,.7))

va_spawn_lfc <- approxfun(va_enhanced_spawn_lfc$flow_cfs, va_enhanced_spawn_lfc$suitable_acres,
                          rule = 2)

# VA rear
new_rearing <- va_proj |>
  filter(habitat_type == 'rearing') |>
  select(reach:max_flow_suitability) |>
  mutate(acres = 2.25 + 3) |>
  unique()

new_rearing_acres <- approxfun(x = c(new_rearing$min_flow, new_rearing$target_flow, new_rearing$max_flow),
                               y = c(new_rearing$acres * new_rearing$min_flow_suitability, 
                                     new_rearing$acres * new_rearing$target_flow_suitability,
                                     new_rearing$acres * new_rearing$max_flow_suitability),
                               yleft = 0, rule = 2)

va_enhanced_rearing_lfc_fry <- tibble(
  flow_cfs = rearing_lfc_fry$flow_cfs,
  suitable_acres = existing_rearing_lfc_fry(flow_cfs) + new_rearing_acres(flow_cfs),
  lifestage = 'fry',
  section = 'LFC'
)

va_enhanced_rearing_lfc_juv <- tibble(
  flow_cfs = rearing_lfc_juv$flow_cfs,
  suitable_acres = existing_rearing_lfc_juv(flow_cfs) + new_rearing_acres(flow_cfs),
  lifestage = 'juvenile',
  section = 'LFC'
)

va_rearing_lfc_fry <- approxfun(va_enhanced_rearing_lfc_fry$flow_cfs, 
                                va_enhanced_rearing_lfc_fry$suitable_acres,
                                rule = 2)
va_rearing_lfc_juv <- approxfun(va_enhanced_rearing_lfc_juv$flow_cfs, 
                                va_enhanced_rearing_lfc_juv$suitable_acres,
                                rule = 2)

bind_rows(
  mutate(rearing_lfc_fry, scenario = 'Existing'),
  mutate(va_enhanced_rearing_lfc_fry, scenario = 'HRL')) |> 
  ggplot(aes(flow_cfs, suitable_acres, color = scenario)) +
  geom_line() +
  theme_minimal() +
  labs(y = 'Suitable acres', x = 'Flow (cfs)', title = "Fry Rearing (LFC)") +
  scale_color_manual(values = plot_palette) +
  theme(legend.justification = c(1,0), legend.position = c(.9,.5))

bind_rows(
  mutate(rearing_lfc_juv, scenario = 'Existing'),
  mutate(va_enhanced_rearing_lfc_juv, scenario = 'HRL')) |> 
  ggplot(aes(flow_cfs, suitable_acres, color = scenario)) +
  geom_line() +
  theme_minimal() +
  labs(y = 'Suitable acres', x = 'Flow (cfs)', title = "Juvenile Rearing (LFC)") +
  scale_color_manual(values = plot_palette) +
  theme(legend.justification = c(1,0), legend.position = c(.9,.5))


# VA floodplain rearing
fp <- va_proj |>
  filter(habitat_type == 'floodplain rearing') |> 
  select(reach:max_flow_suitability)

fp_3000 <- fp |>
  filter(min_flow == 3000) |>
  mutate(acres = 550) |>
  select(-reach) |>
  unique()

new_fp_3000_acres <- approxfun(x = c(fp_3000$min_flow, fp_3000$target_flow, fp_3000$max_flow),
                               y = c(fp_3000$acres * fp_3000$min_flow_suitability, 
                                     fp_3000$acres * fp_3000$target_flow_suitability,
                                     fp_3000$acres * fp_3000$max_flow_suitability),
                               yleft = 0, rule = 2)

fp_30000 <- fp |>
  filter(min_flow == 30000) |>
  mutate(acres = 10 + 220 + 25 + 550) |>
  unique()
new_fp_30000_acres <- approxfun(x = c(fp_30000$min_flow, fp_30000$target_flow, fp_30000$max_flow),
                                y = c(fp_30000$acres * fp_30000$min_flow_suitability, 
                                      fp_30000$acres * fp_30000$target_flow_suitability,
                                      fp_30000$acres * fp_30000$max_flow_suitability),
                                yleft = 0, rule = 2)

fp_4000 <- fp |>
  filter(acres == '100-600') |>
  separate(acres, c('min_acres', 'max_acres'), sep = '-') |>
  mutate(min_acres = 50,
         max_acres = 300,
         min_suit_acres = as.numeric(min_acres) * min_flow_suitability,
         max_suit_acres = as.numeric(max_acres) * max_flow_suitability)
new_fp_4000_acres <- approxfun(x = c(fp_4000$min_flow, fp_4000$target_flow),
                               y = c(fp_4000$min_suit_acres, fp_4000$max_suit_acres),
                               yleft = 0, rule = 2)

flows <- existing_fp$flow_cfs

va_enhanced_fp <- tibble(
  flow_cfs = flows,
  suitable_acres = existing_fp$suitable_acres + new_fp_4000_acres(flows) + 
    new_fp_30000_acres(flows) + new_fp_3000_acres(flows)
)

va_only_fp <- tibble(
  flow_cfs = flows,
  suitable_acres = new_fp_4000_acres(flows) + 
    new_fp_30000_acres(flows) + new_fp_3000_acres(flows)
)

va_fp <- approxfun(va_enhanced_fp$flow_cfs, va_enhanced_fp$suitable_acres, rule = 2)

bind_rows(
  mutate(existing_fp, scenario = 'Existing'),
  mutate(va_enhanced_fp, scenario = 'HRL')) |> 
  ggplot(aes(flow_cfs, suitable_acres, color = scenario)) +
  geom_line() +
  theme_minimal() +
  labs(y = 'Suitable acres', x = 'Flow (cfs)', title = 'Floodplain rearing') +
  scale_color_manual(values = plot_palette) +
  theme(legend.justification = c(1,0), legend.position = c(.9,.3))
```


```{r, echo=FALSE, message = FALSE, warning = FALSE}

feather_spawning <- bind_rows(lfc_spawning,
                              hfc_spawning) |> 
  rename(habitat_area_acres = suitable_acres,
         location = section) |> 
  mutate(watershed = "Feather River",
         habitat_type = "Spawning",
         scenario = "Reference Condition",
         method = "flow-habitat relationship provided",
         interpolation_applied = F) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, habitat_area_acres, method, interpolation_applied)

flows_provided <- c(650,850,1100)
feather_spawn_va <- va_enhanced_spawn_lfc |> 
  rename(habitat_area_acres = suitable_acres) |> 
  mutate(location = "LFC",
         watershed = "Feather River",
         habitat_type = "Spawning",
         scenario = "VA",
         method = "flow-habitat relationship provided",
         interpolation_applied = ifelse(flow_cfs %in% flows_provided, F, T)) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, habitat_area_acres, method, interpolation_applied)

feather_spawn_va_only <- new_spawning |> 
  pivot_longer(cols = c("min_flow", "target_flow",
                        "max_flow"), values_to = "flow_cfs") |> 
  mutate(habitat_area_acres = case_when(name == "min_flow" ~ acres*min_flow_suitability,
                                        name == "target_flow" ~ acres*target_flow_suitability,
                                        name == "max_flow" ~ acres*max_flow_suitability)) |> 
  mutate(location = "LFC",
         watershed = "Feather River",
         habitat_type = "Spawning",
         scenario = "VA additional acres",
         method = "flow-habitat relationship provided",
         interpolation_applied = F) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, habitat_area_acres, method, interpolation_applied)

## instream ----
feather_rearing <- bind_rows(rearing_hfc,
                             rearing_lfc) |> 
  rename(habitat_area_acres = suitable_acres,
         location = section) |> 
  mutate(watershed = "Feather River",
         habitat_type = "Instream rearing",
         scenario = "Reference Condition",
         method = "flow-habitat relationship provided",
         interpolation_applied = F) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, habitat_area_acres, method, interpolation_applied)

flows_provided <- c(650, 725, 3000)
feather_rear_va <- va_enhanced_rearing_lfc_juv |> 
  rename(habitat_area_acres = suitable_acres) |> 
  mutate(location = "LFC",
         watershed = "Feather River",
         habitat_type = "Instream rearing",
         scenario = "VA",
         method = "flow-habitat relationship provided",
         interpolation_applied = ifelse(flow_cfs %in% flows_provided, F, T)) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, habitat_area_acres, method, interpolation_applied)

feather_rear_va_only <- new_rearing |> 
  pivot_longer(cols = c("min_flow", "target_flow",
                        "max_flow"), values_to = "flow_cfs") |> 
  mutate(habitat_area_acres = case_when(name == "min_flow" ~ acres*min_flow_suitability,
                                        name == "target_flow" ~ acres*target_flow_suitability,
                                        name == "max_flow" ~ acres*max_flow_suitability)) |> 
  mutate(location = "LFC",
         watershed = "Feather River",
         habitat_type = "Instream rearing",
         scenario = "VA additional acres",
         method = "flow-habitat relationship provided",
         interpolation_applied = F) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, habitat_area_acres, method, interpolation_applied)

## floodplain ----
feather_existing_fp <- existing_fp |> 
  rename(habitat_area_acres = suitable_acres) |> 
  mutate(location = "Feather River",
         watershed = "Feather River",
         habitat_type = "Floodplain rearing",
         scenario = "Reference Condition",
         method = "flow-habitat relationship provided",
         interpolation_applied = F) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, habitat_area_acres, method, interpolation_applied)

flows_provided <- c(3000,4000, 10000,30000,35000,50000,25000)
feather_va_enhanced_fp <- va_enhanced_fp |> 
  rename(habitat_area_acres = suitable_acres) |> 
  mutate(location = "Feather River",
         watershed = "Feather River",
         habitat_type = "Floodplain rearing",
         scenario = "VA",
         method = "flow-habitat relationship provided",
         interpolation_applied = ifelse(flow_cfs %in% flows_provided, F, T)) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, habitat_area_acres, method, interpolation_applied)

feather_fp_va_only <- bind_rows(select(fp_3000, -habitat_type) |> 
                                  mutate(project_type = "inundation 3000"),
                                select(fp_4000, -c(reach,habitat_type, min_acres, max_acres, min_suit_acres, max_suit_acres)) |> 
                                  mutate(acres = 300,
                                         project_type = "inundation 4000"),
                                select(fp_30000, -c(reach, habitat_type)) |> 
                                  mutate(project_type = "inundation 30000")) |> 
  pivot_longer(cols = 2:4, names_to = "type", values_to = "flow_cfs") |> 
  mutate(habitat_area_acres = case_when(type == "min_flow" ~ acres*min_flow_suitability,
                                        type == "target_flow" ~ acres*target_flow_suitability,
                                        type == "max_flow" ~ acres*max_flow_suitability),
         location = "Feather River",
         watershed = "Feather River",
         habitat_type = "Floodplain rearing",
         scenario = "VA additional acres",
         method = "flow-habitat relationship provided",
         interpolation_applied = F) |> 
  select(watershed, location, scenario, flow_cfs, habitat_type, project_type, habitat_area_acres, method, interpolation_applied) |> 
  distinct()

feather_hrl <- bind_rows(feather_spawning,
                         feather_spawn_va,
                         feather_spawn_va_only,
                         feather_rearing,
                         feather_rear_va,
                         feather_rear_va_only,
                         feather_existing_fp,
                         feather_va_enhanced_fp,
                         feather_fp_va_only)

knitr::kable(head(feather_hrl |> 
                    mutate(scenario = case_when(scenario == "Reference Condition" ~ "Existing",
                                                scenario == "VA" ~ "HRL",
                                                scenario == "VA additional acres" ~ "HRL additional acres")) |> 
                    select(Watershed = watershed, 
                           Location = location,
                           Scenario = scenario,
                           `Flow (cfs)` = flow_cfs,
                           `Habitat type` = habitat_type,
                           `Suitable acres` = habitat_area_acres)))
```

## Mokelumne River

### HRL flow-to-habitat relationship

Instream and floodplain rearing habitat were provided. They are the same as those used in the SBR.

**Data Source:** Data were provided by Robyn Bilski for the SBR.

```{r, echo=FALSE, message = FALSE, warning = FALSE}
spawning <- read_excel('data-raw/R2R_HRL_habitat_inputs/data/moke_flow_to_habitat_RB_edits.xlsx', 
                           sheet = 'spawning', 
                           range = 'A1:D5') |> 
  mutate(FR_spawn_acres_VA_enhanced = FR_spawn_acres_VA_enhanced + FR_spawn_acres_existing,
         FR_sqm_hrl = DSMhabitat::acres_to_square_meters(FR_spawn_acres_VA_enhanced),
         lifestage = "Spawning") |> 
  select(flow_cfs, FR_sqm_hrl, lifestage)

rearing <- read_excel('data-raw/R2R_HRL_habitat_inputs/data/moke_flow_to_habitat_RB_edits.xlsx', 
                      sheet = 'instream_rearing', 
                      range = 'A1:G17') |> 
  mutate(FR_juv_acres_VAadditional = 0.5*(FR_fry_acres_VA_enhanced),
         FR_fry_acres_VA_enhanced = 0.5*(FR_fry_acres_VA_enhanced) + FR_fry_acres_existing,
         FR_juv_acres_VA_enhanced = 0.5*(FR_juv_acres_VA_enhanced) + FR_juv_acres_existing,
         FR_juv_sqm_hrl = acres_to_square_meters(FR_juv_acres_VA_enhanced),
         FR_fry_sqm_hrl = acres_to_square_meters(FR_fry_acres_VA_enhanced)) |> 
  pivot_longer(FR_juv_sqm_hrl:FR_fry_sqm_hrl,
               names_to = "lifestage",
               values_to = "FR_sqm_hrl") |> 
  select(flow_cfs, FR_sqm_hrl, lifestage) 

mokelumne_hrl <- bind_rows(spawning, rearing) |> 
  mutate(scenario = "HRL",
         watershed = "Mokelumne") |> 
  pivot_wider(id_cols = flow_cfs,
              names_from = lifestage,
              values_from = FR_sqm_hrl) |> 
  rename(FR_spw_sqm_hrl = Spawning)

knitr::kable(head(mokelumne_hrl |> 
                    mutate(FR_spw = round(square_meters_to_acres(FR_spw_sqm_hrl)),
                           FR_juv = round(square_meters_to_acres(FR_juv_sqm_hrl)),
                           FR_fry = round(square_meters_to_acres(FR_fry_sqm_hrl))) |> 
                    pivot_longer(FR_spw:FR_fry,
                                 names_to = "habitat_type",
                                 values_to = "suitable_acres") |> 
                    mutate(habitat_type = case_when(habitat_type == "FR_spw" ~ "Spawning",
                                                    habitat_type == "FR_juv" ~ "Juvenile",
                                                    habitat_type == "FR_fry" ~ "Fry"),
                           Scenario = "HRL") |> 
                    select(`Flow (cfs)` = flow_cfs,
                           `Habitat type` = habitat_type,
                           `Suitable acres` = suitable_acres,
                           Scenario)))
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
mokelumne_hrl |> 
  pivot_longer(FR_spw_sqm_hrl:FR_fry_sqm_hrl,
               names_to = "lifestage",
               values_to = "value") |> 
  mutate(lifestage = case_when(lifestage == "FR_fry_sqm_hrl" ~ "Fry",
                               lifestage == "FR_juv_sqm_hrl" ~ "Juvenile",
                               TRUE ~ "Spawning")) |> 
  filter(!is.na(value)) |> 
  mutate(acres = square_meters_to_acres(value)) |> 
  ggplot(aes(x = flow_cfs, y = acres)) + 
  geom_line(color = "#D95F02") +
  facet_wrap(~lifestage, scales = "free") +
  theme_minimal() +
  labs(y = "Suitable acres",
       x = "Flow (cfs)")
```

# Generate estimated habitat using flow

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# american
american_juv_fn <- approxfun(american_hrl$flow_cfs,
                             american_hrl$FR_juv_sqm_hrl,
                             rule = 2)

# feather
# we only need LFC VA and HFC existing


# mokelumne
mokelumne_spw_fn <- approxfun(mokelumne_hrl$flow_cfs,
                              mokelumne_hrl$FR_spw_sqm_hrl,
                              rule = 2)
mokelumne_juv_fn <- approxfun(mokelumne_hrl$flow_cfs,
                              mokelumne_hrl$FR_juv_sqm_hrl,
                              rule = 2)
# TODO not doing fry for mokelumne

# tuolumne
tuolumne_hrl <- read_csv(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", "tuolumne_hrl.csv")) |> 
  mutate(fp_sqm_va = DSMhabitat::acres_to_square_meters(fp_acres_va))

tuolumne_spw_fn <- approxfun(tuolumne_hrl$flow_cfs,
                             tuolumne_hrl$fr_spw_sqm_va,
                             rule = 2)
tuolumne_juv_fn <- approxfun(tuolumne_hrl$flow_cfs,
                             tuolumne_hrl$fr_juv_sqm_va,
                             rule = 2)
tuolumne_fp_fn <- approxfun(tuolumne_hrl$flow_cfs,
                             tuolumne_hrl$fp_sqm_va,
                             rule = 2)

# yuba
yuba_hrl <- read_csv(here::here("data-raw", "R2R_HRL_habitat_inputs", "data", "yuba_hrl.csv"))

yuba_juv_fn <- approxfun(yuba_hrl$flow_cfs,
                         yuba_hrl$fr_juv_ic_sqm,
                         rule = 2)
yuba_fp_fn <- approxfun(yuba_hrl$flow_cfs,
                        yuba_hrl$fr_juv_fp_sqm,
                        rule = 2)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
flows <- DSMflow::flows_cfs$LTO_12a |>
# flows <- DSMflow::flows_cfs$biop_itp_2018_2019 |>
  filter(year(date) >= 1979 & year(date) <= 2000) |> 
  pivot_longer(`Antelope Creek`:`San Joaquin River`,
               names_to = "watershed",
               values_to = "flow_cfs") |>
  mutate(month = month(date),
         year = year(date))

update_hrl_habitat_from_SBR <- function(flows, watershed_name, hab_fn,
                                        is_rearing) {
  
  if(is_rearing) {
    filtered_flows <- flows |> 
      filter(year != 1979)
  } else {
    filtered_flows <- flows
  }

  new_hab <- filtered_flows |>
    filter(watershed == watershed_name) |>
    mutate(new_hab = hab_fn(flow_cfs)) |>
    pivot_wider(id_cols = month, 
                names_from = year, 
                values_from = new_hab) |>
    select(-month) |> 
    as.matrix()

  rownames(new_hab) <- month.abb

  return(new_hab)

}

# update spawning
# start with the baseline habitat run with HRL flows
fr_spawn_update <- DSMhabitat::fr_spawn$r_to_r_lto_12a_baseline

# for feather, add together spawning acres lfc and existing acres hfc
fr_spawn_update["Feather River", , ] <- DSMhabitat::acres_to_square_meters(update_hrl_habitat_from_SBR(flows, "Feather River",
                                                                    va_spawn_lfc, FALSE) + update_hrl_habitat_from_SBR(flows, "Feather River",
                                                                    existing_spawn_hfc, FALSE))

fr_spawn_update["Mokelumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Mokelumne River",
                                                                                    mokelumne_spw_fn, FALSE)
fr_spawn_update["Tuolumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Tuolumne River",
                                                                                    tuolumne_spw_fn, FALSE)
fr_spawn <- DSMhabitat::fr_spawn
fr_spawn$r_to_r_hrl <- fr_spawn_update

# update ic rearing
fr_juv_update <- DSMhabitat::fr_juv$r_to_r_lto_12a_baseline
fr_juv_update["American River", , ] <- update_hrl_habitat_from_SBR(flows, "American River",
                                                                                    american_juv_fn, TRUE)
# feather add together existing hfc and VA lfc
fr_juv_update["Feather River", , ] <- DSMhabitat::acres_to_square_meters(update_hrl_habitat_from_SBR(flows, "Feather River",
                                                                                                     va_rearing_lfc_juv, TRUE) + update_hrl_habitat_from_SBR(flows, "Feather River",
                                                                                                                                            existing_rearing_hfc_juv, TRUE))

fr_juv_update["Mokelumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Mokelumne River",
                                                                                    mokelumne_juv_fn, TRUE)
fr_juv_update["Tuolumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Tuolumne River",
                                                                                    tuolumne_juv_fn, TRUE)
fr_juv_update["Yuba River", , ] <- update_hrl_habitat_from_SBR(flows, "Yuba River",
                                                                                    yuba_juv_fn, TRUE)
fr_juv <- DSMhabitat::fr_juv
fr_juv$r_to_r_hrl <- fr_juv_update

# update fp rearing
fr_fp_update <- DSMhabitat::fr_fp$r_to_r_lto_12a_baseline
# feather fp function by itself is fine, not distinguished by HFC and LFC
fr_fp_update["Feather River", , ] <- DSMhabitat::acres_to_square_meters(update_hrl_habitat_from_SBR(flows, "Feather River",
                                                                                    va_fp, TRUE))

fr_fp_update["Tuolumne River", , ] <- update_hrl_habitat_from_SBR(flows, "Tuolumne River",
                                                                                    tuolumne_fp_fn, TRUE)
fr_fp_update["Yuba River", , ] <- update_hrl_habitat_from_SBR(flows, "Yuba River",
                                                                                    yuba_fp_fn, TRUE)
fr_fp <- DSMhabitat::fr_fp
fr_fp$r_to_r_hrl <- fr_fp_update
```

Habitat is generated using the functionalized flow-to-area curves. Flow inputs are provided by a CalSim3 `LTO_12a` run, which incorporates some HRL flow commitments. The plot below shows an example comparison of juvenile and floodplain habitat (Baseline and HRL) generated using the `LTO_12a` flow run.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# spawning
DSMhabitat::fr_juv$r_to_r_baseline["Tuolumne River",,] |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year", 
               values_to = "sqm") |> 
  mutate(scenario = "Existing") |> 
  bind_rows(fr_juv_update["Tuolumne River",,] |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year", 
               values_to = "sqm") |> 
  mutate(scenario = "HRL")) |>
  mutate(acres = DSMhabitat::square_meters_to_acres(sqm),
         date = as.Date(paste0("1970-", month, "-01"), 
                        format = "%Y-%b-%d")) |> 
  ggplot(aes(x = date, y = acres, color = scenario)) + 
  geom_line() +
  facet_wrap(~year) +
  scale_x_date(labels = date_format("%b")) +
  labs(x = "Model Date",
       y = "Suitable acres",
       title = "Tuolumne River habitat inputs: baseline (LTO flows) vs. HRL (LTO flows)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_manual(values = plot_palette)
```


# Other HRL commitments

We modeled additional commitments that were listed in the [Memorandum of Understanding, Appendix 2](https://resources.ca.gov/-/media/CNRA-Website/Files/NewsRoom/email-items/VoluntaryAgreementMOUTermSheet20220329_SIGNED-20220811.pdf):

| Area                                        | Total acres                   |
|---------------------------------------------|-------------------------------|
| Sutter Bypass, Butte Sink, and Colusa Basin | 20,000 (floodplain)           |
| Sutter Bypass, Butte Sink, and Colusa Basin | 20,000 (fish food production) |

## Sacramento Floodplain improvements (Sutter, Colusa, Butte Sink)

- 20,000 acres were added to the Sutter Bypass watershed **after** the existing flow-to-area relationship was modeled using HRL flows. 
- From December 1st to March 15th, Sutter Bypass will be assumed to have 20,000 acres of suitable habitat in each year.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
va_sutter_habitat_pre <- DSMhabitat::sutter_habitat$lto_12a |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year",
               values_to = "sqm") |> 
  mutate(acres = DSMhabitat::square_meters_to_acres(sqm),
         total_acres = ifelse(month %in% c("Dec", "Jan", "Feb") & 
                                acres < 20000, 20000, acres),
         sqm = DSMhabitat::acres_to_square_meters(total_acres)) |> 
  pivot_wider(id_cols = month, 
              values_from = sqm,
              names_from = year) |> 
  select(-c(month)) |> 
  as.matrix() 

rownames(va_sutter_habitat_pre) <- month.abb

# plot
va_sutter_habitat_pre |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year",
               values_to = "sqm") |> 
  mutate(acres = DSMhabitat::square_meters_to_acres(sqm),
         date = as.Date(paste0(year, "-", month, "-01"), format = "%Y-%b-%d")) |> 
  ggplot(aes(x = date, y = acres)) + 
  geom_line(color = "#D95F02") +
  theme_minimal() +
  labs(x = "Model date",
       y = "Floodplain habitat (acres)",
       title = "Sutter Basin floodplain")
  
```


## 20,000 acres fish food production

- To model the effects of 20,000 acres of fish food production, we calculated the available acreage for dry side floodplain for Sutter and Colusa basins using [these data](https://urldefense.proofpoint.com/v2/url?u=https-3A__flowwest.shinyapps.io_riceflows4ff_&d=DwMGaQ&c=euGZstcaTDllvimEN8b7jXrwqOf-v5A_CdpgnVfiiMM&r=es1mYD_t8ssn_HUUEGOgPpeBoWf2_Iri3BC8scPyjqg&m=MPH46l1pSkkFUhwOq2afwCu7FQy9x-3ynNLgjT1FrFVENIGYvlHv77LDW2IA1HMx&s=gks_3V6uaP9BPa4q9vmFCDSgHkke0BCTa3QXxOITYfw&e=)
- We then calculated the proportion of the committed 20,000 acres those acreages provide, and use that proportion to scale instream habitat for those watersheds.
- This will occur in every year from Jan 1st to Feb 28th

```{r, echo = FALSE, warning = FALSE, message = FALSE}
watersheds <- 
  riceflows4ff::ff_watersheds |> 
  mutate(watershed_group = case_when(
    watershed_name %in% c("Sutter Basin - Upper East", 
                          "Sutter Basin - Lower East",
                          "Gilsizer Slough-Snake River - Upper",
                          "Gilsizer Slough-Snake River - Lower") ~ "Sutter Basin",
    watershed_name %in% c("Colusa Basin Drainage Canal",
                          "Colusa Trough - Lower", 
                          "Colusa Trough - Middle",
                          "Colusa Trough - Upper",
                          "Sycamore Slough - Upper West", 
                          "Freshwater Creek",
                          "Stone Corral Creek", 
                          "Logan Creek", 
                          "Willow Creek", 
                          "Walker Creek",
                          "Colusa Drain") ~ "Colusa Basin")) |>
  filter(!is.na(watershed_group)) |>
  group_by(watershed_group) |>
  summarize() |>
  st_union(by_feature = TRUE)

wet_dry_areas <- 
  watersheds |>
  st_intersection(riceflows4ff::ff_wetdry) |>
  mutate(area_ac = st_area(geometry) |> 
           units::set_units("acres") |>
           units::drop_units()) |>
  group_by(watershed_group, wet_dry) |>
  summarize(area_ac = sum(area_ac)) |>
  ungroup() |>
  st_drop_geometry() |>
  pivot_wider(names_from = wet_dry, values_from = area_ac) |> 
  mutate(proportion = 20000 / Dry)

knitr::kable(head(wet_dry_areas |> 
                    mutate(Proportion = round(proportion, 3),
                           Dry = prettyNum(round(Dry), big.mark = ","),
                           Wet = prettyNum(round(Wet), big.mark = ",")) |> 
                    rename(Watershed = watershed_group) |> 
                    select(-proportion)))
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
va_sutter_habitat_pre <- DSMhabitat::sutter_habitat$lto_12a |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year",
               values_to = "sqm") |> 
  mutate(acres = DSMhabitat::square_meters_to_acres(sqm),
         scenario = "Existing")

va_sutter_habitat_post <- DSMhabitat::sutter_habitat$r_to_r_hrl |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year",
               values_to = "sqm") |> 
  mutate(acres = DSMhabitat::square_meters_to_acres(sqm),
         scenario = "HRL")

# plot
va_sutter_habitat_pre |> 
  bind_rows(va_sutter_habitat_post) |> 
  mutate(date = as.Date(paste0(year, "-", month, "-01"), format = "%Y-%b-%d")) |> 
  ggplot(aes(x = date, y = acres, color = scenario)) + 
  geom_line() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(y = "Floodplain habitat (acres)", 
       x = "Model year",
       title = "Sutter Basin habitat") +
  scale_color_manual(values = plot_palette)
  
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
va_sac_fp_habitat_pre <- DSMhabitat::fr_fp$lto_12a["Lower-mid Sacramento River",,] |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year",
               values_to = "sqm") |> 
  bind_rows(DSMhabitat::fr_fp$lto_12a["Lower Sacramento River",,] |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year",
               values_to = "sqm")) |> 
  mutate(acres = DSMhabitat::square_meters_to_acres(sqm),
         scenario = "Existing")

va_sac_fp_habitat_post <- DSMhabitat::fr_fp$r_to_r_hrl["Lower-mid Sacramento River",,] |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year",
               values_to = "sqm") |> 
  bind_rows(DSMhabitat::fr_fp$r_to_r_hrl["Lower Sacramento River",,] |> 
  as_tibble() |> 
  mutate(month = month.abb) |> 
  pivot_longer(`1980`:`2000`,
               names_to = "year",
               values_to = "sqm")) |> 
  mutate(acres = DSMhabitat::square_meters_to_acres(sqm),
         scenario = "HRL")

# plot
va_sac_fp_habitat_pre |> 
  bind_rows(va_sac_fp_habitat_post) |> 
  mutate(date = as.Date(paste0(year, "-", month, "-01"), format = "%Y-%b-%d")) |> 
  ggplot(aes(x = date, y = acres, color = scenario)) + 
  geom_line() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(y = "Floodplain habitat (acres)", 
       x = "Model year",
       title = "Sacramento floodplain (Colusa basin and Butte sink) habitat") +
  scale_color_manual(values = plot_palette)
  
```
```
