---
title: "American River HRL curves"
author: "[Liz Stebbins](mailto:lstebbins@flowwest.com)" 
date:  "July 2024" 
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, out.width = '100%')
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(usethis.quiet = TRUE)
library(tidyverse)
library(readxl)
library(DSMhabitat)
library(scales)
```

## Future Data Improvements

American River will have updated existing habitat curves in fall of 2024.

## HRL flow-to-habitat relationship

**Data Source:** FlowWest coordinated with representatives for the American River to determine the best scientific information available to quantify existing suitable habitat, and to describe the VA non-flow habitat as *suitable area that varies with flow*. Habitat data were provided to R2R for use in accounting for planned habitat projects by [Erica Bishop](mailto:ebishop@waterforum.org) and [Chris Hammersmark](mailto:c.hammersmark@cbecoeng.com). Data provided represented the HRL commitments based on flow-to-habitat relationships borne out by recent EIP designs and constructed projects for the subject habitat type. 

```{r, echo=FALSE}
american_hrl <- read_excel("data-raw/R2R_HRL_habitat_inputs/data/HRL_habitat_commitments_for_R2R_American.xlsx",
                               range = "A1:G10") |> 
  janitor::clean_names() |> 
  mutate(duration_inundated_days = ifelse(duration_inundated_days == "N/A", NA, duration_inundated_days),
         water_year_type = ifelse(water_year_type == "N/A", NA, water_year_type),
         run = ifelse(run == "fall run", "fall", run)) |> 
  mutate(flow_cfs = as.numeric(flow_to_inundate_suitable_acres),
         flow_cfs = ifelse(is.na(flow_cfs), 20000, flow_cfs), 
         scenario = "HRL", 
         watershed = "American River") |> 
  select(watershed, flow_cfs, FR_juv_wua_hrl = suitable_acres, scenario)

knitr::kable(head(american_hrl))
```

### Plot comparison

- flows are different for existing and HRL commitments, so to compare we make a functionalized version of the existing curve and use it to map to the flows for the HRL commitments.

- is floodplain cutoff `5200 cfs`?

```{r, echo = FALSE, warning = FALSE, message = FALSE}
existing_rearing <- DSMhabitat::american_river_instream |> 
  filter(!is.na(FR_juv_wua)) |> 
  select(flow_cfs, FR_juv_wua)

existing_floodplain <- DSMhabitat::american_river_floodplain |> 
  filter(!is.na(FR_floodplain_acres)) |> 
  select(flow_cfs, FR_floodplain_acres)

existing_rearing_fn <- approxfun(existing_rearing$flow_cfs, existing_rearing$FR_juv_wua, rule = 2)

existing_fp_fn <- approxfun(existing_floodplain$flow_cfs, existing_floodplain$FR_floodplain_acres, rule = 2)

existing_and_HRL <- american_hrl |> 
  mutate(existing_rear = existing_rearing_fn(flow_cfs),
         existing_fp = existing_fp_fn(flow_cfs),
         existing_and_HRL = ifelse(flow_cfs <= 5200, existing_rear + FR_juv_wua_hrl, existing_fp + FR_juv_wua_hrl)) |> 
  select(-scenario)

existing_and_HRL |> 
  pivot_longer(c(existing_rear, existing_fp, existing_and_HRL),
               names_to = "scenario",
               values_to = "value") |> 
  ggplot(aes(x = flow_cfs, y = value)) + 
  geom_line() +
  facet_wrap(~scenario, scales = "free_y") +
  geom_vline(aes(xintercept = 5200), linetype = "dashed") +
  theme_minimal()

#usethis::use_data(american_river_instream, overwrite = TRUE)
```
