---
title: "Tuolumne River HRL curves"
author: "[Liz Stebbins](mailto:lstebbins@flowwest.com)" 
date:  "August 2024" 
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, out.width = '100%')
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(usethis.quiet = TRUE)
library(tidyverse)
library(readxl)
library(DSMhabitat)
library(scales)
```

## Future Data Improvements

None expected.

## HRL flow-to-habitat relationship

**Data Source:** Existing and existing + VA habitat curves for the Tuolumne river were provided by William Sears and Tuolumne parties. The TUFlow modeling report from the FERC relicensing, which describes the process for developing the 1D/2D surface from which rasters of depth and velocity were derived for the entire 50 mile reach of the lower Tuolumne, can be accessed [here](https://va-hrl-docs.s3.amazonaws.com/tuolumne/Tuolumne+River+VA+Habitat+Modeling+for+SBR+final+revision+20240802.pdf).


```{r, echo=FALSE}
spawning_habitat_updated <- read_excel("data-raw/watershed/tuolumne/data/VA habitat summary 20240509[44][35].xlsx",
                                       range = "A6:C34") |> 
  janitor::clean_names() |> 
  mutate(fr_spw_sqm_existing = DSMhabitat::acres_to_square_meters(existing),
         fr_spw_sqm_va = DSMhabitat::acres_to_square_meters(existing_va)) |> 
  select(flow_cfs = la_grange_flow_cfs, fr_spw_sqm_existing, fr_spw_sqm_va)

rearing_habitat_updated <- read_excel("data-raw/watershed/tuolumne/data/VA habitat summary 20240509[44][35].xlsx",
                                      range = "E6:G53") |> 
  janitor::clean_names() |> 
  mutate(fr_juv_sqm_existing = DSMhabitat::acres_to_square_meters(existing),
         fr_juv_sqm_va = DSMhabitat::acres_to_square_meters(existing_va)) |> 
  select(flow_cfs, fr_juv_sqm_existing, fr_juv_sqm_va)

# spawning and rearing in this updated file are the same as in the file
# from 2024-05-10. This file just has "wetted" habitat
wetted_area <- read_excel("data-raw/watershed/tuolumne/data/VA habitat summary 20240603.xlsx",
                          range = "I6:K53") |> 
  janitor::clean_names() |> 
  mutate(fp_acres_existing = existing,
         fp_acres_va = existing_va) |> 
  select(flow_cfs, fp_acres_existing, fp_acres_va)

tuolumne_hrl <- full_join(spawning_habitat_updated, 
                          rearing_habitat_updated,
                          by = "flow_cfs") |> 
  full_join(wetted_area)

knitr::kable(head(tuolumne_hrl))
```

### Plot comparison

```{r, echo = FALSE, warning = FALSE, message = FALSE}
tuolumne_hrl |> 
  pivot_longer(c(fr_spw_sqm_existing:fp_acres_va),
               names_to = "scenario",
               values_to = "value") |> 
  mutate(lifestage = case_when(str_detect(scenario, "spw") ~ "spawn",
                               str_detect(scenario, "juv") ~ "juv",
                               str_detect(scenario, "fp") ~ "floodplain"),
         scenario = ifelse(str_detect(scenario, "va"), "VA", "existing")) |> 
  ggplot(aes(x = flow_cfs, y = value, color = scenario)) + 
  geom_line() +
  facet_wrap(~lifestage, scales = "free_y") +
  theme_minimal()

write_csv(tuolumne_hrl, here::here("data-raw", "R2R_HRL_habitat_inputs",
                                     "data", "tuolumne_hrl.csv"))
```
