---
title: "CVPIA Habitat Values"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output:
  rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE, 
                      message = FALSE)

library(tidyverse)
library(DSMhabitat)
library(lubridate)
```

```{r include=FALSE}
# load functions: 

roll_stat = min
annual_stat = max

exceedance_probs_monthly <- function(data, roll_stat, annual_stat) {
  annual_durations <- data %>%
    mutate(water_year = ifelse(month(date) %in% 10:12, year(date) + 1, year(date))) 
  
  annual_stats <- annual_durations %>%
    group_by(water_year) %>%
    summarise(stat_in_duration = annual_stat(flow_cfs, na.rm = TRUE))
  
  annual_stats %>%
    mutate(dist = round(cume_dist(-stat_in_duration), 3)) %>%
    arrange(dist)
  
}

calsim_30_day <- function(data, exceedance_function) {
  dur_30_min_max <- exceedance_probs_monthly(data, roll_stat = roll_stat, annual_stat = annual_stat)
  
  interpolate_probs_30_min_max <- approxfun(x = dur_30_min_max$dist, 
                                            dur_30_min_max$stat_in_duration)
  
  d30 <- interpolate_probs_30_min_max(0.5) 
  
  return(d30)
}


existing_flow_cfs <- function(habitat_type, watershed_input, bypass = FALSE) {

  flow_df <- if (bypass == FALSE) {DSMflow::flows_cfs$biop_itp_2018_2019} else {DSMflow::bypass_flows$biop_itp_2018_2019}
  
  if(habitat_type == "spawning") {
    flow_df |> 
      filter(date >= as_date("1980-01-01")) |> 
      filter(month(date) %in% spawning_months) |> 
      pull(watershed_input) |> 
      median()
  } else if(habitat_type == "rearing") {
    flow_df |> 
      filter(date >= as_date("1980-01-01")) |> 
      filter(month(date) %in% rearing_months) |> 
      pull(watershed_input) |> 
      median()
  } else if(habitat_type == "flood") {
    flood = flow_df |> 
      filter(date >= as_date("1980-01-01")) |> 
      select(watershed_input, date) |> 
      rename(flow_cfs = watershed_input) 
    
    exceedance_probs_monthly_fun <- exceedance_probs_monthly(flood, roll_stat, annual_stat)
    calsim_30_day(flood, exceedance_probs_monthly_fun)
    
  }
}

```

## Flood Duration Methodology

**Objective**: Calculate the flow that is exceeded in 50% of the years 

**Method**: Take the yearly rolling maximum of monthly CalSim flows > 1980. Apply a cumulative distribution function on the ranked values to find the proportion of all values less than or equal to the ranked order. Create an interpolation function to find the 2 year flow value

```{r}
watershed_input <- "Cosumnes River"

flood_flow_ex = DSMflow::flows_cfs$biop_itp_2018_2019 |> 
      filter(date >= as_date("1980-01-01")) |> 
      select(watershed_input, date) |> 
      rename(flow_cfs = watershed_input) 
    
exceedance_probs_monthly_fun <- exceedance_probs_monthly(flood_flow_ex, 'min', 'max')
calsim_30_day(flood_flow_ex, exceedance_probs_monthly_fun)

ggplot(exceedance_probs_monthly_fun, aes(x = dist, y = stat_in_duration)) + 
  geom_line() + 
  ggtitle('option 3 interpolation example: want 50% distance value')
    
    
# plot it: 
ggplot(DSMflow::flows_cfs$biop_itp_2018_2019 |> 
      filter(date >= as_date("1980-01-01")) |> 
      select(watershed_input, date) |> 
      rename(flow_cfs = watershed_input)) +
  geom_line(aes(x = date, y = flow_cfs)) +
  geom_hline(aes(yintercept = calsim_30_day(flood_flow_ex, exceedance_probs_monthly_fun), color = "2 year flow"))

```

## Calculate Habitat for SIT Watersheds

#### Define Inputs: 

```{r}
# missing sutter bypass and yolo bypass 
watersheds_labels <- DSMscenario::watershed_labels[-c(17, 22)]


# not included in floodplain suitability factor 
modeling_in_suitable_area <- c("Antelope Creek", "Battle Creek", "Bear Creek", 
                               "Cow Creek", "Mill Creek", "Paynes Creek", 
                               "Deer Creek",'Upper Sacramento River',
                               'Upper-mid Sacramento River','Lower Sacramento River')

spawning_months <- c(10:12)
rearing_months <- c(1:8)

```

```{r}
sit_outputs <- data.frame()
for(i in 1:length(watersheds_labels)) {
  watershed_input <- watersheds_labels[i]
  
  if (watershed_input == "Lower-mid Sacramento River") {
    # The Lower-mid Sacramento River has two nodes, one above Fremont Weir (C134) and one below (C160). 
    # rearing: 
    rear_flow1 <- existing_flow_cfs("rearing", "Lower-mid Sacramento River1")
    rear_flow2 <- existing_flow_cfs("rearing", "Lower-mid Sacramento River2")
    
    rear_acres_juv <- square_meters_to_acres(DSMhabitat::set_instream_habitat('Lower-mid Sacramento River', 
                                                                              species = "fr", life_stage = "juv",
                                                                              rear_flow1, rear_flow2))
    rear_acres_fry <- square_meters_to_acres(DSMhabitat::set_instream_habitat('Lower-mid Sacramento River',
                                                                              "fr", "fry",  rear_flow1, rear_flow2))
    
    # floodplain: 
    flood_flow1 <- existing_flow_cfs("flood", "Lower-mid Sacramento River1")
    flood_flow2 <- existing_flow_cfs("flood", "Lower-mid Sacramento River2")
    flow_acres = square_meters_to_acres(set_floodplain_habitat("Lower-mid Sacramento River", "fr", 
                                                               flood_flow1, flood_flow2))
    

    spwn_acres = NA
    spwn_flow = NA
    rear_flow = NA
    flood_flow = NA
    
  } else {
    spwn_flow <- existing_flow_cfs("spawning", watershed_input)
    spwn_acres <- square_meters_to_acres(set_spawning_habitat(watershed_input, "fr", spwn_flow, month = 10))
    
    rear_flow <- existing_flow_cfs("rearing", watershed_input)
    rear_acres_juv <- square_meters_to_acres(DSMhabitat::set_instream_habitat(watershed_input, "fr", "juv", rear_flow))
    rear_acres_fry <- square_meters_to_acres(DSMhabitat::set_instream_habitat(watershed_input, "fr", "fry", rear_flow))
    
    flood_flow <- existing_flow_cfs("flood", watershed_input)
    flood_acres <- square_meters_to_acres(DSMhabitat::set_floodplain_habitat(watershed_input, "fr", flood_flow))
  }
  
  if (!(watershed_input %in% modeling_in_suitable_area)) {
    flood_acres <- DSMhabitat::apply_suitability(flood_acres)
  }
  
  sit_outputs <- bind_rows(sit_outputs, 
                           tibble(
                             watershed = watershed_input, 
                             spwn_flow = spwn_flow, 
                             spwn_acres = spwn_acres, 
                             rear_flow = rear_flow, 
                             rear_acres_juv = rear_acres_juv,
                             rear_acres_fry = rear_acres_fry,
                             flood_flow = flood_flow, 
                             flood_acres = flood_acres
                           )
  )
  
}

knitr::kable(sit_outputs, digits = 1)

write_csv(sit_outputs,'../CVPIA_habitat_data.csv')

```

## Calculate Bypass Habitat: 

Only suitable for rearing? https://cvpia-osc.github.io/DSMhabitat/reference/bypass_habitat.html

```{r}
bypasses <- tibble(bypass = c(rep("Yolo Bypass",2), rep("Sutter Bypass",4 )),
                   names = c("yolo1", "yolo2", "sutter1", "sutter2", "sutter3", "sutter4"))

bypass_outputs <- data.frame()
for(i in 1:length(bypasses$names)) {
  bypass_outputs <- bind_rows(bypass_outputs, 
                              tibble(
                                bypass_name = bypasses$names[i], 
                                bypass = bypasses$bypass[i],
                                rear_flow = existing_flow_cfs("rearing", bypasses$names[i], bypass = TRUE),
                                rear_habitat = square_meters_to_acres(DSMhabitat::set_bypass_habitat(bypasses$names[i], rear_flow))
                              )
  )
  
}

bypass_totals <- bypass_outputs |> 
  group_by(bypass) |> 
  summarise(total_rear_acres = sum(rear_habitat))

knitr::kable(bypass_totals, digits = 1)
```


## Delta Habitats
```{r}
delta_rearing_habitat |> 
  select(date, `North Delta`) |> 
  filter(month(date) %in% rearing_months) |> 
  summarise(mean_area = mean(`North Delta`),
            mean_acres = square_meters_to_acres(mean_area),
            median = square_meters_to_acres(median(`North Delta`))) |> 
  knitr::kable()

delta_rearing_habitat |> 
  select(date, `South Delta`) |> 
  filter(month(date) %in% rearing_months) |> 
  summarise(mean_area = mean(`South Delta`),
            mean_acres = square_meters_to_acres(mean_area)) |> 
  knitr::kable()
```