---
title: "Deer Creek"
author: "[Sadie Gill](mailto:sgill@flowwest.com), [Mark Tompkins](mailto:mtompkins@flowwest.com)" 
date: "June 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, out.width = '100%')
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(usethis.quiet = TRUE)
library(tidyverse)
library(readxl)
library(DSMhabitat)
library(lubridate)
```   


```{r}
source('data-raw/watershed/floodplain_utils.R')
```

```{r}
# the rearing lengths need to scaled, based on the model length and the actual rearing
# the sr spawning needs to be approximated using the regional method
rearing_extents <- DSMhabitat::watershed_lengths %>% 
  filter(watershed == "Deer Creek", lifestage == "rearing") %>% 
  select(species, miles) %>% 
  spread(species, miles)

# there is a timestep that contains 250 cfs as well, I have
# made the decision to remove timestep: 2018-11-15 11:00:00
raw_data <- read_csv("data-raw/watershed/deer/data/existing-suitability-deer-creek-2021-07-14.csv") %>% 
  filter(timestamp != as_datetime("2018-11-15 11:00:00"))

sr_scale <- (rearing_extents$sr - rearing_extents$fr) / rearing_extents$fr

st_scale <- (rearing_extents$st - rearing_extents$fr) / rearing_extents$fr

# get all data in appropriate names
deer_creek_habitat <- raw_data %>% 
  filter(species %in% c("fall", "late-fall", "spring", "steelhead"), 
         life_stage %in% c("fry", "rearing", "spawning")) %>%
  mutate(suitable_acres = ifelse(suitable_acres < 0, NA_real_, suitable_acres)) %>% 
  transmute(
    flow_cfs = flow, 
    sqm = acres_to_square_meters(suitable_acres), 
    suitable_acres,
    species = case_when(
      species == "fall" ~ "fr", 
      species == "spring" ~ "sr", 
      species == "late-fall" ~ "lfr", 
      species == "steelhead" ~ "st", # double check
      TRUE ~ NA_character_
    ),
    life_stage = case_when(
      life_stage == "adult" ~ "adult", 
      life_stage == "rearing" ~ "juv", 
      life_stage == "spawning" ~ "spawn", 
      life_stage == "fry" ~ "fry"
    ), 
    habitat_type, 
    watershed = "Deer Creek"
  )
```

## Instream Spawning and Rearing Habitat
**Data Source:** [FlowWest 2021](https://cvpiahabitat-r-package.s3.us-west-2.amazonaws.com/cvpia-sit-model-inputs/DeerCreek_2Dmodel_FlowWest_Final.pdf){target="_blank"}

A flow to suitable area relationship was modeled using depth and velocity suitability ranges applied to HEC-RAS 2-dimensional hydraulic model output for the instream spawning habitat of Fall Run Chinook Salmon, Late-Fall Run Chinook Salmon and the instream rearing habitat of Fall Run Chinook Salmon, Late-Fall Run Chinook Salmon, Spring Run Chinook Salmon, and Steelhead, as well as fry habitat for Fall Run Chinook Salmon, Late-Fall Run Chinook Salmon, Spring Run Chinook Salmon, and Steelhead. Spring Run Chinook and Steelhead spawning habitat were [calculated with a relationship between flow and mean weighted usable area from watersheds in the region with modeled habitat](http://cvpia-habitat-docs-markdown.s3-website-us-west-2.amazonaws.com/watershed/Regional_Approximation.html){target="_blank"}.

Suitable ranges for Deer Creek were defined using the following sources:

* Fall Run & Late-Fall Run Spawning - Clear Creek Fall Run Suitability Criteria (Gard 2019)
* All Species Juvenile Instream Rearing - Clear Creek Fall Run Suitability Criteria (Gard 2019)
* All Species Juvenile Floodplain Rearing - CBEC 2019
* Fall, Spring & Late-Fall Fry Rearing - Battle Creek Fall Run Suitability Criteria (Gard 2019)
* Steelhead Fry Rearing - Battle Creek Steelhead Suitability Criteria (Gard 2019)

Data from Gard 2019 can be downloaded [here](https://cvpia-documents.s3.us-west-1.amazonaws.com/HSC.xlsx).

The depth and velocity ranges for species and life-stage combinations are listed in the following table. 

| **Species** | **Lifestage** | **Location** | **Suitable Depth Range (feet)** | **Suitable Velocity Range (feet)** | 
  | --- | --- | --- | --- | --- | --- |
  | Fall Run Chinook, Late-Fall Run Chinook | Adult Spawning | Instream | 0.41 - 6.69 | 0.09 - 6.30 | 
  | Fall Run Chinook, Late-Fall Run Chinook | Juvenile Rearing | Instream | 0.41 - 5.31 | 0.00 - 3.07 |
  | Spring Run Chinook, Steelhead | Juvenile Rearing | Instream | 0.21 - 5.59 | 0.00 - 5.54 | 
  | Fall Run Chinook, Late-Fall Run Chinook, Spring Run Chinook, Steelhead | Juvenile Rearing | Floodplain | 0.50 - 5.20 | 0.00 - 4.00 | 
  | Fall Run Chinook, Late-Fall Run Chinook, Spring Run Chinook | Fry | Instream | 0.05 - 3.45 | 0 - 1.82 | 
  | Steelhead | Fry | Instream | 0.09 - 4 | 0.09 - 3.6 | 
  
```{r}
# instream
# we need to scale sr_juv and st_juv only
# sr_spawn needs to be regional approximated
deer_creek_instream <-
  deer_creek_habitat %>% 
  filter(habitat_type == "instream") %>% 
  transmute(
    flow_cfs, 
    col_to_spread = sprintf("%s_%s_sqm", toupper(species), life_stage), 
    sqm, 
    watershed
  ) %>% 
    spread(col_to_spread, sqm) %>% 
  transmute(
    flow_cfs, 
    FR_spawn_sqm, 
    FR_juv_sqm,
    FR_fry_sqm,
    LFR_spawn_sqm, 
    LFR_juv_sqm,
    LFR_fry_sqm,
    SR_juv_sqm = (SR_juv_sqm + ((sr_scale*SR_juv_sqm)*.1)),
    SR_fry_sqm = (SR_fry_sqm + ((sr_scale*SR_fry_sqm)*.1)),
    ST_juv_sqm = (ST_juv_sqm + ((st_scale*ST_juv_sqm)*.1)),
    ST_fry_sqm = (ST_fry_sqm + ((st_scale*ST_fry_sqm)*.1)),
    watershed
  )

usethis::use_data(deer_creek_instream, overwrite = TRUE)
```

### Spawning Data
```{r}
spawn_table <- deer_creek_instream %>% select(flow_cfs, contains("spawn"))
knitr::kable(align = 'c', head(spawn_table, 5), 
             caption = "Header Descriptions: flow_cfs = flow in cubic feet per second,
             FR_spawn_sqm = Fall Run Chinook Spawning Area (square meters),
             LFR_spawn_sqm = Late-Fall Run Chinook Spawning Area (square meters), 
             watershed = section of stream modeled for CVPIA SDM")
```
*... with `r nrow(deer_creek_instream) - 5` more rows*


### Rearing Data

```{r}
rearing_table <- deer_creek_instream %>% select(flow_cfs, contains(c("juv","fry")))
knitr::kable(align = 'c', head(rearing_table, 5), 
             caption = "Header Descriptions: flow_cfs = flow in cubic feet per second,
             FR_juv_sqm = Fall Run Chinook Juvenile Area (square meters), 
             LFR_juv_sqm = Late-Fall Run Chinook Spawning Area (square meters), 
             SR_juv_sqm = Spring Run Chinook Juvenile Area (square meters),
             ST_juv_sqm = Steelhead Juvenile Area (square meters),
             FR_fry_sqm = Fall Run Chinook Juvenile Area (square meters), 
             LFR_fry_sqm = Late-Fall Run Chinook Spawning Area (square meters), 
             SR_fry_sqm = Spring Run Chinook Juvenile Area (square meters),
             ST_fry_sqm = Steelhead Juvenile Area (square meters)")
```
*... with `r nrow(deer_creek_instream) - 5` more rows*



### Spawning Habitat Area 
The following plot shows the suitable spawning habitat in acres for Fall Run and
Late-Fall Run Chinook. 

```{r}
deer_creek_instream %>% 
  select(flow_cfs, `Fall and Late-Fall Run Chinook` = FR_spawn_sqm) %>% 
  gather(species, sqm, -flow_cfs) %>% 
  ggplot(aes(flow_cfs, sqm/4046.86, color = species)) + 
  geom_line() + 
  labs(x = 'Flow (cfs)', y = 'Suitable Habitat (acres)', color = NULL) + 
  theme_minimal() + 
  scale_color_brewer(palette = 'Dark2') +
  theme(legend.justification = c(1, 0), legend.position = c(.97, .7))
```

### Fry Rearing Habitat Area

The following plot shows the suitable fry habitat in acres for Fall Run Chinook,
Late-Fall Run Chinook, Spring Run Chinook and Steelhead.

```{r}
deer_creek_instream %>% 
  select(flow_cfs, `Fall and Late-Fall Chinook` = FR_fry_sqm, 
         `Spring Run Chinook` = SR_fry_sqm,
         `Steelhead` = ST_fry_sqm) %>% 
  gather(species, sqm, -flow_cfs) %>% 
  ggplot(aes(flow_cfs, sqm/4047, color = species)) + 
  geom_line() + 
  labs(x = 'Flow (cfs)', y = "Suitable Habitat (acres)", color = NULL) + 
  theme_minimal() + 
  scale_color_brewer(palette = 'Dark2') +
  theme(legend.justification = c(1, 0), legend.position = c(.97, .7))

```


### Juvenile Rearing Habitat Area 
The following plot shows the suitable rearing habitat in acres for Fall Run Chinook, 
Late-Fall Run Chinook, Spring Run Chinook and Steelhead. 
```{r}
deer_creek_instream %>% 
  select(flow_cfs, `Fall and Late-Fall Run Chinook` = FR_juv_sqm, 
         `Spring Run Chinook and Steelhead` = SR_juv_sqm) %>% 
  gather(species, sqm, -flow_cfs) %>% 
  ggplot(aes(flow_cfs, sqm/4046.86, color = species)) + 
  geom_line() + 
  labs(x = 'Flow (cfs)', y = "Suitable Habitat (acres)", color = NULL) + 
  theme_minimal() + 
  scale_color_brewer(palette = 'Dark2') +
  theme(legend.justification = c(1, 0), legend.position = c(.97, .7))
```


## Floodplain Rearing Habitat 

**Data Source:** [Calcuated using a flow to area relationship modeled using HEC-RAS (FlowWest 2021)](https://cvpiahabitat-r-package.s3.us-west-2.amazonaws.com/cvpia-sit-model-inputs/DeerCreek_2Dmodel_FlowWest_Final.pdf){target="_blank"}

### Fall Run and Late-Fall Run
`r print_model_details('Deer Creek', 'fr')`


### Spring Run and Steelhead 
A 12.2 mile portion of the entire mapped 
rearing extent of 47.9 miles was modeled using Lower Deer Creek Flood and Ecosystem Improvement Project HEC-RAS Hydrodynamic model. 
Of the entire mapped rearing extent, 12.2 miles were classified 
as low gradient and 35.7 miles were classified as high gradient 
based on a geomorphic analysis of long profile slopes and valley widths. 
The floodplain area per unit length was determined for the modeled 
extent and used to approximate areas for the non-modeled extent. 
The area per unit length was scaled by a factor of 0.1 
for the high gradient extent. There was no scaling factor applied 
to the low gradient extent.

## Floodplain Data    
The areas represent total suitable rearing area in acres.

```{r}
deer_creek_floodplain <- 
  deer_creek_habitat %>% 
  filter(habitat_type == "floodplain", life_stage == "juv") %>% 
  transmute(
    flow_cfs, 
    col_to_spread = sprintf("%s_floodplain_acres", toupper(species)), 
    suitable_acres,
    watershed
  ) %>% 
  spread(col_to_spread, suitable_acres) %>% 
  transmute(
    flow_cfs, 
    FR_floodplain_acres, 
    LFR_floodplain_acres, 
    SR_floodplain_acres = (SR_floodplain_acres + (sr_scale*SR_floodplain_acres)*.1),
    ST_floodplain_acres = (ST_floodplain_acres + (st_scale*ST_floodplain_acres)*.1),
    watershed
  ) %>% 
  relocate(watershed, .after = ST_floodplain_acres)

knitr::kable(align = 'c', head(deer_creek_floodplain, 5), 
             caption = "Header Descriptions: flow_cfs = flow in cubic feet per second,
             FR_floodplain_acres = Fall Run Chinook floodplain acres, SR_floodplain_acres 
             = Spring Run Chinook floodplain acres, ST_floodplain_acres = 
             Steelhead floodplain acres, 
             watershed = section of stream modeled for CVPIA SDM")

usethis::use_data(deer_creek_floodplain, overwrite = TRUE)
```

*... with `r nrow(deer_creek_floodplain) - 5` more rows*

## Floodplain Plot
```{r}
deer_creek_floodplain %>% 
  select(`Fall and Late-Fall Run Chinook` = FR_floodplain_acres, 
         `Spring Run Chinook and Steelhead` = SR_floodplain_acres, flow_cfs) %>%
  gather(species, acres, -flow_cfs) %>% 
  ggplot(aes(flow_cfs, acres, color = species)) +
  geom_line() +
  theme_minimal() +
  scale_color_brewer(palette = 'Dark2') +
  labs(x = 'Flow (cfs)', y = 'Total Suitable Area (acres)') +
  theme(legend.justification = c(1,0), legend.position = c(1,.1))
```
