---
title: "Cottonwood Creek"
author: "[Sadie Gill](mailto:sgill@flowwest.com), [Mark Tompkins](mailto:mtompkins@flowwest.com)"
date: "July 2020"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, out.width = '100%')
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(usethis.quiet = TRUE)
library(tidyverse)
library(readxl)
library(DSMhabitat)
library(scales)
```
## Future Data Improvements
None planned. 

## Instream Spawning and Rearing Habitat
**Data Source:** [CDFW 1979](https://s3-us-west-2.amazonaws.com/cvpiahabitat-r-package/cvpia-sit-model-inputs/CDFW1979_CottonwoodSpawningIFIM.PDF){target="_blank"} (pgs 11-15) and [UFWS 2013](https://cvpiahabitat-r-package.s3-us-west-2.amazonaws.com/cvpia-sit-model-inputs/CVPIA_Annual_Progress_Report_Fiscal_Year_2013.pdf){target="_blank"} (pgs 46-47)

The instream spawning area for Fall Run Chinook Salmon in Cottonwood Creek is based on data from a collaborative study in 1979 by the U.S. Army Corps of Engineers and the California Department of Fish and Wildlife. The federal instream flow method was used to measure the rearing habitat at three reaches, that together span a 22.4 mile stretch of Cottonwood Creek. Spawning habitat data were provided on pages 11-15 (Table 1 and Table 5) of the CDFW 1979 report. Instream rearing habitat for Fall Run Chinook Salmon and Steelhead are based on the UFWS 2013 study that included 26 habitat transects on lower Cottonwood Creek. The habitat data from this study are on pages 46 and 47 of the USFWS 2014 report. The instream spawning and rearing habitat data were provided to us by [Mark Gard](mailto:Mark.Gard@Wildlife.ca.gov) from the California Department of Fish and Wildlife.

### Spawning and Rearing Data    
The instream spawning and rearing habitat data described above for Fall Run Chinook Salmon (FR) is combined for use in the Decision Support Model (DSM) in the following format. Units are in square feet per 1000 feet.
```{r}
lens <- read_excel('data-raw/watershed/cottonwood/data/cottonwood_combined.xlsx', 
                   sheet = "reach_lengths") %>% 
  spread(reach, miles)

total_length <- lens$`Cottonwood Creek Downstream of South Fork` + lens$`Cottonwood Creek Upstream of South Fork`

# spawning
cottonwood_raw <- read_excel('data-raw/watershed/cottonwood/data/cottonwood_combined.xlsx', sheet = "spawning_both_reaches") 

spawn_reach_1 <- cottonwood_raw %>% 
  filter(reach == 1)

spawn_reach_3 <- cottonwood_raw %>% 
  filter(reach == 3)

flows <- unique(cottonwood_raw$Flow)

cotton_spawn_values_reach_1 <- tibble(
  flow = flows,
  Spawning = approx(spawn_reach_1$Flow, spawn_reach_1$Spawning, 
                    xout = flows, rule = 2)$y,
  miles = lens$`Cottonwood Creek Downstream of South Fork`
)

cotton_spawn_values_reach_3 <- tibble(
  flow = flows,
  Spawning = approx(spawn_reach_3$Flow, spawn_reach_3$Spawning, 
                      xout = flows, rule = 2)$y,
  miles = lens$`Cottonwood Creek Upstream of South Fork`
)

cottonwood_spawning <- cotton_spawn_values_reach_1 %>%
  bind_rows(cotton_spawn_values_reach_3) %>%
  gather(species_lifestage, wua, -flow, -miles) %>%
  mutate(reach_proportion = miles/total_length,
         scaled_wua = wua * reach_proportion) %>%
  group_by(flow, species_lifestage) %>%
  summarise(wua = sum(scaled_wua)) %>%
  ungroup() %>%
  mutate(FR_spawn_wua = wua/total_length/5.28) %>%
  select(flow_cfs = flow, FR_spawn_wua)


# Rearing
downstream_south_fork <- read_excel('data-raw/watershed/cottonwood/data/cottonwood_combined.xlsx', 
                                    sheet = "downstream_of_south_fork") %>% 
  mutate(miles = lens$`Cottonwood Creek Downstream of South Fork`)

upstream_south_fork <- read_excel('data-raw/watershed/cottonwood/data/cottonwood_combined.xlsx', 
                                    sheet = "upstream_of_south_fork") %>% 
  mutate(miles = lens$`Cottonwood Creek Upstream of South Fork`)

new_cottonwood_creek_instream <- bind_rows(downstream_south_fork, upstream_south_fork) %>% 
  gather(species_lifestage, wua, -Flow, -miles) %>% 
  mutate(reach_proportion = miles/total_length,
         scaled_wua = wua * reach_proportion) %>% 
  group_by(Flow, species_lifestage) %>% 
  summarise(wua = sum(scaled_wua)) %>% 
  ungroup() %>% 
  mutate(wua = wua/total_length/5.28) %>% 
  spread(species_lifestage, wua) %>% 
  arrange(Flow) %>% 
   rename(flow_cfs = Flow, FR_fry_wua = `FR fry`, 
          FR_juv_wua = `FR Juv`, ST_fry_wua = `ST fry`, ST_juv_wua = `ST juv`) %>% 
   mutate(watershed = 'Cottonwood Creek')

cottonwood_creek_instream <- cottonwood_spawning %>% 
  full_join(new_cottonwood_creek_instream) %>%
  arrange(flow_cfs) %>%
  mutate(watershed = 'Cottonwood Creek')

usethis::use_data(cottonwood_creek_instream, overwrite = TRUE)
```
```{r}
knitr::kable(align = 'c', head(cottonwood_creek_instream, 5), 
             caption = "Header Descriptions: flow_cfs = flow in cubic feet per second,
             FR_spawn_wua = Fall Run Chinook Spawning WUA, FR_fry_wua = 
             Fall Run Chinook Fry WUA, FR_juv_wua = Fall Run Chinook Juvenile WUA, 
             ST_fry_wua = Steelhead Fry WUA, ST_juv_wua = Steelhead Juvenile WUA,
             watershed = section of stream modeled for CVPIA SDM")
```

*... with `r nrow(cottonwood_creek_instream) - 5` more rows*

### Spawning WUA 
The following plot shows the weighted usable spawning area (WUA) in square feet per thousand feet for Fall Run Chinook Salmon. These area per length rates are multiplied by the total spawning reach length mapped by the Science Integration Team (SIT) to estimate available habitat. 
``` {r}
cottonwood_creek_instream %>% 
    filter(!is.na(FR_spawn_wua)) %>%
    ggplot(aes(x = flow_cfs , y = FR_spawn_wua)) +
    geom_line(color = '#7570b3') +
    labs(x = 'Flow (cfs)', y = 'WUA (sqft/1000ft)') + 
    theme_minimal()
```

### Rearing WUA
The following plot shows the rearing WUA in square feet per thousand feet for Fall Run Chinook Salmon and Steelhead fry and juvenile. These rates are multiplied by the total rearing reach length mapped by the SIT to estimate available habitat. 
``` {r}
cottonwood_creek_instream %>% 
  select(FR_juv_wua, FR_fry_wua, ST_juv_wua, ST_fry_wua, flow_cfs, watershed) %>%
  gather(species_lifestage, acres, -flow_cfs, -watershed) %>% 
  mutate(
    species = case_when( 
      str_detect(species_lifestage, 'ST') ~ 'Steelhead', 
      str_detect(species_lifestage, 'FR') ~ 'Fall Run Chinook'),
      lifestage = ifelse(str_detect(species_lifestage, 'juv'), 'Juvenile', 'Fry')) %>% 
  ggplot(aes(flow_cfs, acres, color = species)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~lifestage, ncol = 1, scales = 'free_y') +
  scale_color_brewer(palette = 'Dark2') +
  labs(x = 'Flow (cfs)', y = 'WUA (sqft/1000ft)') + 
  theme(legend.justification = c(1,0), legend.position = c(.95,.7))
```

```{r}
source('data-raw/watershed/floodplain_utils.R')
```

## Floodplain Rearing Habitat
**Data Source:** [USFWS / FEMA 1D HEC-RAS hydraulic model](https://s3-us-west-2.amazonaws.com/cvpiahabitat-r-package/cvpia-sit-model-inputs/CVPIA+Annual+Progress+Report+Fiscal+Year+2017.pdf){target="_blank"}

### Fall Run
`r print_model_details('Cottonwood Creek', 'fr')`

### Spring Run
`r print_model_details('Cottonwood Creek', 'sr')`

### Steelhead
`r print_model_details('Cottonwood Creek', 'st')`

## Floodplain Data   
The areas represent total inundated area in acres.    
```{r}
# Using 49 acres wetted channel area that mark gard developed because google earth channel area is 250 too big
df <- read_excel('data-raw/mark_gard_data/Floodplainarea.xlsx', 
                            sheet = "Cottonwood", skip = 2, 
                            col_names = c('trash', 'flow_cfs', 'modeled_floodplain_area_acres', 
                                          'total_area_acres', 'trash3')) %>% 
  select(flow_cfs, modeled_floodplain_area_acres)

cottonwood_creek_floodplain <- scale_fp_flow_area_partial_model(ws = "Cottonwood Creek", df)

knitr::kable(align = 'c', head(cottonwood_creek_floodplain, 5), 
             caption = "Header Descriptions: flow_cfs = flow in cubic feet per second,
             FR_floodplain_acres = Fall Run Chinook floodplain acres, 
             SR_floodplain_acres = Spring Run Chinook floodplain acres, 
             ST_floodplain_acres = Steelhead floodplain acres,
             watershed = section of stream modeled for CVPIA SDM")

usethis::use_data(cottonwood_creek_floodplain, overwrite = TRUE)
```

*... with `r nrow(cottonwood_creek_floodplain) - 5` more rows*

## Floodplain Plot
```{r}
cottonwood_creek_floodplain %>% 
  rename(`Fall Run Chinook` = FR_floodplain_acres,
         `Spring Run Chinook` = SR_floodplain_acres,
         `Steelhead` = ST_floodplain_acres) %>% 
  gather(Species, acres, -flow_cfs, -watershed) %>% 
  ggplot(aes(flow_cfs, acres, color = Species)) +
  geom_line() +
  theme_minimal() +
  scale_color_brewer(palette = 'Dark2') +
  labs(x = 'Flow (cfs)', y = 'Total Inundated Acres') +
  theme(legend.justification = c(1,0), legend.position = c(1,.1))
```
