---
title: "Tuolumne River"
author: "[Sadie Gill](mailto:sgill@flowwest.com), [Mark Tompkins](mailto:mtompkins@flowwest.com)" 
date: "July 2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, out.width = '100%')
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(usethis.quiet = TRUE)
library(tidyverse)
library(readxl)
library(DSMhabitat)
```

## Future Data Improvement
None expected.

## Instream Spawning and Rearing Habitat
**Data Source:** [Methods for Quantifying Existing and Future Suitable Habitat for Fall-run Chinook Salmon, 2024](https://va-hrl-docs.s3.amazonaws.com/tuolumne/Tuolumne+River+VA+Habitat+Modeling+for+SBR+final+revision+20240802.pdf){target="_blank"}

Instream spawning and rearing habitat for Fall Run Chinook Salmon in the Tuolumne River were provided by
William Sears and Tuolumne parties as part of the Healthy Rivers and Landscapes modeling effort. The TUFlow modeling report from the FERC relicensing, which describes the process for developing the 1D/2D surface from which rasters of depth and velocity were derived for the entire 50 mile reach of the lower Tuolumne River. These were provided in units of suitable acres.

### Steelhead
The instream spawning and rearing habitat for Steelhead in the Tuolumne River are based on data from a California Department of Fish and Game site-specific study. These data consist of 10 sites spread over 9.2 miles of the dominant spawning reach of the Lower Tuolumne River. They were reviewed by Stillwater Sciences for use in their 2014 Flo2D model. The weighted usable area (WUA) results for Steelhead are provided in tables H-2 on pages 108-109 of the [Final Lower Tuolumne Instream Flow Study Report](https://s3-us-west-2.amazonaws.com/cvpiahabitat-r-package/cvpia-sit-model-inputs/Stillwater2010_Tuolumne_P-2299-075_71_DP_FLA_AttC_StudyRept_IFIM_AppA-H_140428.pdf){target="_blank}. The instream spawning and rearing habitat data were provided to us by [Mark Gard](mailto:Mark.Gard@Wildlife.ca.gov) from the California Department of Fish and Wildlife in a [spreadsheet](https://cvpiahabitat-r-package.s3-us-west-2.amazonaws.com/IFIMWUA.xlsx). Mark Gard instructed us to use cells A2:H32 in the 'Tuolumne' tab within the spreadsheet.

### Spawning and Rearing Data 
The instream spawning and rearing habitat data described above for Fall Run Chinook Salmon (FR) were used for Steelhead (ST). Units are in suitable acres and converted to square meters.
```{r}
tuolumne_river_instream_rearing <- read_excel("data-raw/watershed/tuolumne/data/VA habitat summary 20240509[44][35].xlsx",
                                      range = "E6:G53") |> 
  janitor::clean_names() |> 
  mutate(FR_fry_sqm = DSMhabitat::acres_to_square_meters(existing),
         FR_juv_sqm = DSMhabitat::acres_to_square_meters(existing)) |> 
  select(flow_cfs, FR_fry_sqm, FR_juv_sqm)

tuolumne_river_spawning <- read_excel("data-raw/watershed/tuolumne/data/VA habitat summary 20240509[44][35].xlsx",
                                       range = "A6:C34") |> 
  janitor::clean_names() |> 
  mutate(FR_spawn_sqm = DSMhabitat::acres_to_square_meters(existing)) |> 
  select(flow_cfs = la_grange_flow_cfs, FR_spawn_sqm)

tuolumne_river_instream <- full_join(tuolumne_river_instream_rearing, tuolumne_river_spawning,
                                     by = "flow_cfs") |> 
  mutate(ST_adult_sqm = FR_spawn_sqm,
         ST_juv_sqm = FR_juv_sqm,
         ST_fry_sqm = FR_fry_sqm,
         watershed = "Tuolumne River")

usethis::use_data(tuolumne_river_instream, overwrite = TRUE)

knitr::kable(align = 'c', head(tuolumne_river_instream, 5), 
             caption = "Header Descriptions: flow_cfs = flow in cubic feet per second,
             FR_fry_sqm = Fall Run Chinook fry (square meters), 
             FR_juv_sqm = Fall Run Chinook Juvenile (square meters), 
             FR_spawn_sqm = Fall Run Chinook Spawning (square meters), 
             ST_adult_sqm = Steelhead Adult (square meters),
             ST_juv_sqm = Steelhead Juvenile (square meters), 
             ST_fry_sqm = Steelhead fry (square meters), 
             watershed = section of stream modeled for CVPIA SDM")
```

*... with `r nrow(tuolumne_river_instream) - 5` more rows*

### Spawning WUA 
The following plot shows the spawning suitable area in square meters for Fall Run Chinook Salmon and Steelhead. Steelhead values are assumed to be the same as Fall Run Chinook.

``` {r}
tuolumne_river_instream |> 
    select(flow_cfs, `Fall Run Chinook` = FR_spawn_sqm, `Steelhead` = ST_adult_sqm) |>
    gather(Species, sqm, -flow_cfs) |> 
    filter(!is.na(sqm)) |>
    ggplot(aes(flow_cfs, sqm, color = Species)) +
    geom_line() +
    labs(x = 'Flow (cfs)', y = 'Suitable area (square meters)') + theme_minimal()  +
    scale_color_manual(values = c('#d95f02','#7570b3')) +
    theme(legend.justification = c(1, 0), legend.position = c(1, .5))
```

### Rearing WUA
The following plot shows the rearing WUA in square feet per thousand feet for Fall Run Chinook Salmon and Steelhead fry and juvenile. These rates are multiplied by the total rearing reach length mapped by the SIT to estimate available habitat.
``` {r}
tuolumne_river_instream |> 
  select(FR_juv_sqm, FR_fry_sqm, ST_juv_sqm, ST_fry_sqm, flow_cfs, watershed) |>
  gather(species_lifestage, square_meters, -flow_cfs, -watershed) |> 
  mutate(
    Species = case_when( 
      str_detect(species_lifestage, 'ST') ~ 'Steelhead', 
      str_detect(species_lifestage, 'FR') ~ 'Fall Run'),
      lifestage = ifelse(str_detect(species_lifestage, 'juv'), 'Juvenile', 'Fry')) |> 
  ggplot(aes(flow_cfs, square_meters, color = Species)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~lifestage, ncol = 1, scales = 'free_y') +
  scale_color_brewer(palette = 'Dark2') +
  labs(x = 'Flow (cfs)', y = 'Suitable area (square meters)') + 
  theme(legend.justification = c(1, 0), legend.position = c(.95, .25))
```

```{r, include=FALSE}
source('data-raw/watershed/floodplain_utils.R')
```

## Floodplain Rearing Habitat    

## Floodplain Data    
The areas represent total inundated area in acres.    
```{r}
tuolumne_fp <- read_excel("data-raw/watershed/tuolumne/data/VA habitat summary 20240603.xlsx",
                          range = "I6:K53") |> 
  janitor::clean_names() |> 
  mutate(FR_floodplain_acres = round(existing, 2),
         SR_floodplain_acres = FR_floodplain_acres,
         ST_floodplain_acres = FR_floodplain_acres,
         watershed = "Tuolumne River") |> 
  select(flow_cfs,
         FR_floodplain_acres,
         SR_floodplain_acres, ST_floodplain_acres,
         watershed)

knitr::kable(align = 'c', head(tuolumne_river_floodplain, 5), 
             caption = "Header Descriptions: flow_cfs = flow in cubic feet per second,
             FR_floodplain_acres = Fall Run Chinook floodplain acres, 
             SR_floodplain_acres = Spring Run Chinook floodplain acres, 
             ST_floodplain_acres = Steelhead floodplain acres,
             watershed = section of stream modeled for CVPIA SDM")

usethis::use_data(tuolumne_river_floodplain, overwrite = TRUE)
```

*... with `r nrow(tuolumne_river_floodplain) - 5` more rows*

## Floodplain Plot
```{r}
tuolumne_river_floodplain |> 
  select(FR_floodplain_acres, flow_cfs, watershed) |>
  rename(`Fall Run Chinook,
Spring Run chinook, 
and Steelhead` = FR_floodplain_acres) |> 
  gather(Species, acres, -flow_cfs, -watershed) |> 
  ggplot(aes(flow_cfs, acres, color = Species)) +
  geom_line() +
  theme_minimal() +
  scale_color_brewer(palette = 'Dark2') +
  labs(x = 'Flow (cfs)', y = 'Total Inundated Acres') +
  theme(legend.justification = c(1, 0), legend.position = c(1, .1))
```


